---
title: "PFAS and Proteomics Praxis-BGM Clustering"
author: "Zhenjiang Li"
date: today
toc: true
toc_expand: 2
format:
  html:
    code-fold: true
    html-math-method: katex
    fontsize: 16pt
    grid:
      sidebar-width: 0px
      body-width: 1500px
      margin-width: 500px
      gutter-width: 1.5rem
embed-resources: true
editor: visual
editor_options:
  chunk_output_type: console
---

## Environment Setup {.hidden}

```{=html}
<style type="text/css">
caption, .table-caption {
  text-align: left;
}
</style>
```

```{r setup}
#| echo: false

options(scipen = 999)

library(fs)
library(here)
i_am("1_scripts/!analytic_workflow_pfas_praxis_BGM.qmd")

source(here::here("1_scripts", "!libraries.R"))
source(here::here("1_scripts", "!directories.R"))
source(here::here("1_scripts", "!functions.R"))
source(here::here("1_scripts", "0_reactome_pathway_helpers.R"))

library(ggtext)

options(knitr.table.format = "html")
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r shorten-long-path}
#| include: false

create_shortcut <- function(path, short_cut, drive) {
  return(
    sub(gsub("\\)", "\\\\)", gsub("\\(", "\\\\(", short_cut)), paste0(drive, ":"), path)
  )
}

if (nchar(dir_data) > 80) {
  system("subst x: /D")
  short_cut <- dir_home |> dirname()
  system(paste0("subst x: \"", short_cut, "\""))

  dir_objects_names <- ls(pattern = "^dir_")
  dir_list <- mget(dir_objects_names)

  dir_list_shortened <- lapply(
    dir_list,
    function(y) create_shortcut(y, short_cut, "x")
  )
  list2env(dir_list_shortened, envir = .GlobalEnv)
}

rm(list = intersect(
  c("dir_objects_names", "short_cut", "dir_list", "dir_list_shortened"),
  ls()
))
```

# Parameters

```{r parameters}
cohort_names <- c("SOLAR", "CHS", "BHRC")

SOLAR_pfas_names <- c("pfos_log2", "pfoa_log2", "pfna_log2", "pfda_log2", "pfhxs_log2", "pfhps_log2")
CHS_pfas_names <- c("ma_pfos_log2", "ma_pfoa_log2", "ma_pfna_log2", "ma_pfda_log2", "ma_pfhxs_log2", "ma_pfhps_log2")
BHRC_pfas_names <- c("PFOS_det60_imputed_log2", "PFOA_det60_imputed_log2", "PFNA_det60_imputed_log2", "PFDA_det60_imputed_log2", "PFHxS_det60_imputed_log2", "PFHpS_det60_imputed_log2")
PFAS_abbrev_names <- c("PFOS", "PFOA", "PFNA", "PFDA", "PFHxS", "PFHpS")

praxis_k <- 2
praxis_seed <- 20260716
praxis_num_iters <- 500
praxis_beta <- 0.001
praxis_tol <- 1e-4
praxis_max_iters <- 300L
praxis_prior_mus_variance <- 10.0
praxis_num_samples <- 16
praxis_data_precision_int <- 1L
praxis_likelihood_temp <- 1.0
praxis_rho_prec <- 0.05
praxis_rho_mu <- 1.0
praxis_elbo_eval_freq <- 5L
praxis_batch_size_max <- 64
praxis_early_stop <- TRUE
praxis_patience <- 5L
praxis_covariance_transfer_scale <- 1
pwas_screening_fdr_cutoff <- 0.20
marker_bf_threshold <- 1
pfas_concentration_unit <- "ng/mL"

dir_praxis <- fs::path(dir_temp_data, "Praxis_BGM")
dir_pwas <- fs::path(dir_temp_data, "PFAS_PWAS")
fs::dir_create(dir_praxis)
```

The Praxis-BGM parameters were chosen to match the scientific goal and keep the first-pass analysis computationally stable.

- `praxis_k = 2`: Directly encodes the target of identifying two participant clusters.
- `praxis_seed = 20260716`: Fixes the random initialization so the same data and settings give reproducible cluster assignments.
- `praxis_num_iters = 500`: Uses a moderate first-pass training length, long enough for the model to move beyond initialization but short enough for exploratory runs. If cluster assignments are unstable across seeds or diagnostics suggest non-convergence, increase this value.
- `praxis_beta = 0.001`: Controls the strength of the Praxis-BGM prior/regularization term used by the wrapper. This small value keeps the model weakly regularized rather than forcing clusters too strongly.
- `praxis_tol = 1e-4`: Sets the convergence tolerance used during model initialization.
- `praxis_max_iters = 300`: Caps the number of initialization iterations before the main Praxis-BGM optimization loop.
- `praxis_prior_mus_variance = 10.0`: Sets a diffuse variance for automatically initialized cluster means.
- `praxis_num_samples = 16`: Sets the Monte Carlo/sample count used by Praxis-BGM during fitting and feature scoring, balancing runtime against posterior stability.
- `praxis_data_precision_int = 1`: Passes the data precision control used by the Python implementation.
- `praxis_likelihood_temp = 1.0`: Uses the standard, untempered likelihood contribution.
- `praxis_rho_prec = 0.05`: Applies conservative damping to precision/covariance updates.
- `praxis_rho_mu = 1.0`: Applies full mean updates.
- `praxis_elbo_eval_freq = 5`: Evaluates the ELBO every five iterations for convergence monitoring.
- `praxis_batch_size_max = 64`: Caps the mini-batch size so fitting is feasible for the cohort sample sizes and local machine memory.
- `praxis_early_stop = TRUE` and `praxis_patience = 5`: Stop fitting early after five ELBO checks without sufficient improvement.
- `praxis_covariance_transfer_scale = 1`: Transfers the SOLAR posterior covariance to CHS and BHRC without inflating or shrinking uncertainty. Values above 1 make the transferred prior more diffuse, while values below 1 make it tighter.
- `pwas_screening_fdr_cutoff = 0.20`: Selects proteins associated with any single PFAS compound or the PFAS mixture at BH-adjusted p \< 0.20 in SOLAR.
- `marker_bf_threshold = 1`: Displays proteins with Praxis-BGM Bayes-factor importance values at or above 1 in the marker protein abundance plot.
- `pfas_concentration_unit = "ng/mL"`: Labels the back-transformed PFAS concentrations in the visualization section. Change this if the cohort source data use a different PFAS concentration unit.

# Praxis-BGM Interface

Praxis-BGM is fit from R through the `Praxis_BGM_R_interface` wrapper.
The wrapper uses `reticulate` and expects the Python package `praxis_bgm` to be installed in the Python environment selected below.

```{r praxis-interface}
library(reticulate)

praxis_python <- Sys.getenv("RETICULATE_PYTHON", unset = "")
praxis_python <- Sys.getenv("PRAXIS_PYTHON", unset = praxis_python)

if (!nzchar(praxis_python)) {
  praxis_python <- "C:/Users/lizhenji/AppData/Local/Python/pythoncore-3.12-64/python.exe"
}

reticulate::use_python(praxis_python, required = TRUE)
reticulate::py_config()

praxis_r_interface_path <- Sys.getenv("PRAXIS_BGM_R_INTERFACE", unset = "")

if (nzchar(praxis_r_interface_path) && fs::file_exists(praxis_r_interface_path)) {
  source(praxis_r_interface_path)
} else {
  source("https://raw.githubusercontent.com/BruceResearch/Praxis_BGM_R_interface/main/R/praxis_bgm_interface.R")
}

if (!reticulate::py_module_available("praxis_bgm")) {
  stop(
    "The Python module 'praxis_bgm' is not available. ",
    "Install it in the selected Python environment, for example: ",
    "pip install git+https://github.com/ContiLab-usc/Praxis-BGM.git"
  )
}
```

# Data Preparation

This section harmonizes the three cohorts before clustering.
The goal is to build one participant-level table per cohort with the same six PFAS compounds and the same bridged Olink protein features.
SOLAR is treated as the reference cohort because its fitted Praxis-BGM posterior is later transferred to CHS and BHRC as prior information.

## Bridged proteomics

The bridged proteomics object contains cohort-specific Olink matrices and protein annotation.
We keep only proteins observed in all three cohorts so that the SOLAR reference model and the CHS/BHRC transfer models use the same feature space.
Cohort-specific protein identifiers are then converted to participant IDs used by the PFAS files.

```{r proteomics-data}
proteomics_l <- readRDS(fs::path(
  dir_data,
  "ShARP_BHRC_Shared",
  "ShARP_BHRC_Data_Shared",
  "OLINK_proteomics_bridged_BHRC_SOLAR_CHS.rds"
))

proteomics_metadata_cols <- c("LABID", "study", "Batch")

overlap_proteins <- proteomics_l[cohort_names] %>%
  purrr::map(~ setdiff(names(.x), proteomics_metadata_cols)) %>%
  purrr::reduce(intersect)

proteomics_l <- proteomics_l[c(cohort_names, "protein_annotation")]

proteomics_l[cohort_names] <- proteomics_l[cohort_names] %>%
  purrr::map(~ .x %>% dplyr::select(tidyselect::any_of(proteomics_metadata_cols), tidyselect::all_of(overlap_proteins)))

proteomics_l$protein_annotation <- proteomics_l$protein_annotation[cohort_names] %>%
  purrr::map(~ .x %>% dplyr::filter(meta_id %in% overlap_proteins))

SOLAR_prote <- proteomics_l$SOLAR %>%
  dplyr::mutate(id = gsub("-", "_", tolower(LABID))) %>%
  dplyr::select(id, tidyselect::all_of(overlap_proteins))

CHS_prote <- proteomics_l$CHS %>%
  dplyr::rename(SampleID = LABID) %>%
  dplyr::select(SampleID, tidyselect::all_of(overlap_proteins))

BHRC_prote <- proteomics_l$BHRC %>%
  dplyr::select(LABID, tidyselect::all_of(overlap_proteins))

proteomics_property <- proteomics_l$protein_annotation$SOLAR %>%
  dplyr::transmute(
    OlinkID = meta_id,
    UniProt = meta_Uniprot,
    Assay = meta_id,
    Protein = meta_Assay,
    Panel = NA_character_
  )

cat("Overlapping proteins across SOLAR, CHS, and BHRC:", length(overlap_proteins), "\n")
```

## SOLAR

For SOLAR, the first available visit per participant is used.
PFAS variables are joined to bridged protein abundances through the harmonized participant ID.

```{r solar-data-prep}
load(fs::path(
  dir_data,
  "ShARP_SOLAR_Shared",
  "ShARP_SOLAR_Data_Shared",
  "Solar_Analysis_20251003.Rda"
))
SOLAR <- analysis_df
rm(analysis_df)

SOLAR_visit1 <- SOLAR %>%
  dplyr::group_by(id) %>%
  dplyr::filter(visit == min(visit)) %>%
  dplyr::ungroup() %>%
  dplyr::select(id, tidyselect::all_of(SOLAR_pfas_names))

SOLAR_working_df <- SOLAR_prote %>%
  dplyr::left_join(SOLAR_visit1, by = "id")

dim(SOLAR_working_df)
```

## CHS

For CHS, the MetaAir baseline records provide PFAS concentrations.
The plasma sample inventory links proteomics sample IDs to CHS participant IDs before joining to the analysis data.

```{r chs-data-prep}
CHS <- readRDS(fs::path(
  dir_data,
  "ShARP_CHS_Shared",
  "ShARP_CHS_Data_Shared",
  "CHS MetaAir MetaChem Cleaned Redcap and Exposures Outcome Data long V5.rds"
))

CHS_id_link <- readxl::read_xlsx(
  fs::path(
    here::here() |> dirname() |> dirname() |> dirname() |> dirname() |> dirname(),
    "Chatzi Projects and Analyses",
    "Env Chem SOL-CHS",
    "Lab Analysis and sample shipments",
    "Metachem_sample_inventory_master_040522.xlsx"
  ),
  sheet = "Plasma"
) %>%
  dplyr::transmute(
    SampleID = `Plasma Aliquot ID`,
    id = as.character(`Sample Name`)
  )

CHS_prote <- CHS_prote %>%
  dplyr::left_join(CHS_id_link, by = "SampleID") %>%
  dplyr::select(id, tidyselect::everything(), -SampleID)

CHS_baseline <- CHS %>%
  dplyr::filter(study == "MetaAir") %>%
  dplyr::select(id, tidyselect::all_of(CHS_pfas_names))

CHS_working_df <- CHS_prote %>%
  dplyr::left_join(CHS_baseline, by = "id")

dim(CHS_working_df)
```

## BHRC

For BHRC, visit 1 is used for PFAS and joined to bridged protein abundances by `LABID`.
The PFAS variables here use the detection/imputation fields from the three-set PFAS file to match the PWAS screening workflow.

```{r bhrc-data-prep}
BHRC <- readRDS(
  fs::path(
    dir_data,
    "ShARP_BHRC_Shared",
    "ShARP_BHRC_Data_Shared",
    "CCHC_usc250_long_12172025.rds"
  )
) %>%
  dplyr::filter(visit %in% c(1, 2))

BHRC_pfas <- readRDS(
  fs::path(
    dir_data,
    "ShARP_BHRC_Shared",
    "ShARP_BHRC_Data_Shared",
    "PFAS_CCHC_three_sets.rds"
  )
)

BHRC_analysis_df <- BHRC %>%
  dplyr::filter(visit == 1) %>%
  dplyr::select(RRID, LABID, visit) %>%
  dplyr::left_join(
    BHRC_pfas %>%
      dplyr::select(RRID, tidyselect::all_of(BHRC_pfas_names)),
    by = "RRID"
  ) %>%
  dplyr::arrange(LABID, visit) %>%
  dplyr::distinct(LABID, .keep_all = TRUE)

BHRC_working_df <- BHRC_prote %>%
  dplyr::left_join(BHRC_analysis_df, by = "LABID")

dim(BHRC_working_df)
```

# Protein Features

Praxis-BGM is fit using proteins only in this workflow.
PFAS variables are not included in the model matrix; they are used upstream to screen proteins and downstream to characterize the resulting clusters.
The protein screen uses the PWAS results generated by `!analytic_workflow_pfas_PWAS.qmd`.
Proteins are retained if they are associated with any single PFAS compound or the PFAS mixture at BH-adjusted p \< 0.20 in SOLAR.

## PWAS-Based Protein Selection

```{r pwas-protein-selection}
pwas_single_file <- fs::path(dir_pwas, "PFAS_single_PWAS_results.csv")
pwas_mixture_file <- fs::path(dir_pwas, "PFAS_mixture_qgcomp_PWAS_results.csv")

if (!fs::file_exists(pwas_single_file) || !fs::file_exists(pwas_mixture_file)) {
  stop(
    "PWAS screening files were not found. Run 1_scripts/!analytic_workflow_pfas_PWAS.qmd first."
  )
}

PFAS_single_PWAS_results <- readr::read_csv(pwas_single_file, show_col_types = FALSE) %>%
  dplyr::filter(cohort == "SOLAR")

PFAS_mixture_PWAS_results <- readr::read_csv(pwas_mixture_file, show_col_types = FALSE) %>%
  dplyr::filter(cohort == "SOLAR")

selected_single_pwas_proteins <- PFAS_single_PWAS_results %>%
  dplyr::filter(!is.na(adjusted_pval_BH), adjusted_pval_BH < pwas_screening_fdr_cutoff) %>%
  dplyr::pull(feature_name)

selected_mixture_pwas_proteins <- PFAS_mixture_PWAS_results %>%
  dplyr::filter(!is.na(adjusted_pval_BH), adjusted_pval_BH < pwas_screening_fdr_cutoff) %>%
  dplyr::pull(feature_name)

selected_proteins <- union(selected_single_pwas_proteins, selected_mixture_pwas_proteins) %>%
  intersect(overlap_proteins) %>%
  sort()

if (length(selected_proteins) == 0) {
  stop("No proteins passed the BH-adjusted PWAS/qgcomp screening threshold.")
}

tibble::tibble(
  n_overlap_proteins = length(overlap_proteins),
  n_single_PFAS_selected = length(intersect(unique(selected_single_pwas_proteins), overlap_proteins)),
  n_mixture_selected = length(intersect(unique(selected_mixture_pwas_proteins), overlap_proteins)),
  n_proteins_used_in_praxis = length(selected_proteins),
  protein_filtering = paste0(
    "SOLAR single PFAS or PFAS mixture association with BH-adjusted p < ",
    pwas_screening_fdr_cutoff
  )
) %>%
  kableExtra::kbl(caption = "Praxis-BGM protein feature summary.") %>%
  kableExtra::kable_paper("hover", full_width = FALSE)

PFAS_single_PWAS_results %>%
  dplyr::filter(!is.na(adjusted_pval_BH), adjusted_pval_BH < pwas_screening_fdr_cutoff) %>%
  dplyr::count(cohort, pfas_abbrev, name = "n_selected") %>%
  dplyr::arrange(cohort, pfas_abbrev) %>%
  dplyr::rename(
    `Cohort` = cohort,
    `PFAS` = pfas_abbrev,
    `Proteins selected` = n_selected
  ) %>%
  kableExtra::kbl(caption = "Proteins selected by SOLAR single-PFAS PWAS at BH-adjusted p < 0.20.") %>%
  kableExtra::kable_paper("hover", full_width = FALSE)

pwas_selected_protein_table <- dplyr::bind_rows(
  PFAS_single_PWAS_results %>%
    dplyr::filter(!is.na(adjusted_pval_BH), adjusted_pval_BH < pwas_screening_fdr_cutoff) %>%
    dplyr::transmute(
      analysis = "Single-PFAS PWAS",
      exposure = pfas_abbrev,
      feature_name,
      Protein,
      UniProt,
      estimate = estimate_beta,
      standard_error = se_beta,
      ci_low = ci_beta_low,
      ci_high = ci_beta_high,
      test_statistic,
      p_value,
      adjusted_pval_BH,
      adjusted_pval_BY
    ),
  PFAS_mixture_PWAS_results %>%
    dplyr::filter(!is.na(adjusted_pval_BH), adjusted_pval_BH < pwas_screening_fdr_cutoff) %>%
    dplyr::transmute(
      analysis = "PFAS-mixture qgcomp",
      exposure = "PFAS mixture",
      feature_name,
      Protein,
      UniProt,
      estimate = estimate_beta,
      standard_error = NA_real_,
      ci_low = ci_beta_low,
      ci_high = ci_beta_high,
      test_statistic,
      p_value,
      adjusted_pval_BH,
      adjusted_pval_BY
    )
) %>%
  dplyr::filter(feature_name %in% selected_proteins) %>%
  dplyr::arrange(adjusted_pval_BH, p_value, Protein, exposure)

pwas_selected_protein_table %>%
  dplyr::transmute(
    `Selection analysis` = analysis,
    `Associated exposure` = exposure,
    `Olink assay` = feature_name,
    `Protein name` = Protein,
    `UniProt ID` = UniProt,
    `Effect estimate` = round(estimate, 3),
    `Standard error` = round(standard_error, 3),
    `95% CI` = dplyr::if_else(
      !is.na(ci_low) & !is.na(ci_high),
      paste0("[", round(ci_low, 3), ", ", round(ci_high, 3), "]"),
      NA_character_
    ),
    `Test statistic` = round(test_statistic, 3),
    `P-value` = scales::pvalue(p_value, accuracy = 0.001),
    `BH-adjusted p-value` = scales::pvalue(adjusted_pval_BH, accuracy = 0.001),
    `BY-adjusted p-value` = scales::pvalue(adjusted_pval_BY, accuracy = 0.001)
  ) %>%
  kableExtra::kbl(
    caption = "Supplemental Table. SOLAR PWAS associations used to select proteins for Praxis-BGM.",
    align = c("l", "l", "l", "l", "l", rep("r", 7))
  ) %>%
  kableExtra::kable_paper("hover", full_width = FALSE) %>%
  kableExtra::footnote(
    general = paste(
      "Proteins were selected for Praxis-BGM if they were associated with any single PFAS compound or the PFAS mixture in SOLAR at BH-adjusted p < 0.20.",
      "Effect estimates and 95% confidence intervals are from the SOLAR PWAS screening models.",
      "For single-PFAS PWAS, the effect estimate is the beta coefficient for the listed PFAS exposure.",
      "For PFAS-mixture qgcomp, the effect estimate is the mixture effect estimate.",
      "BH, Benjamini-Hochberg; BY, Benjamini-Yekutieli; CI, confidence interval."
    ),
    general_title = "Note: ",
    footnote_as_chunk = TRUE
  )
```

# Praxis-BGM Data Matrices

The Praxis-BGM input matrix contains only the selected proteins.
PFAS variables are kept in the cohort data for post hoc cluster visualization, but they are not passed to Praxis-BGM.
Protein abundances are already log2-transformed upstream.
The scaling step below does not log-transform again; it z-scores each protein feature using the SOLAR reference mean and SD.
CHS and BHRC are then placed on the same SOLAR reference scale before transfer modeling.

## Matrix construction helpers

```{r praxis-data-helpers}
rename_pfas_to_common_names <- function(df, pfas_names, common_names = PFAS_abbrev_names) {
  out <- df
  names(out)[match(pfas_names, names(out))] <- common_names
  out
}

make_praxis_df <- function(df, id_col, pfas_names, protein_names) {
  df %>%
    dplyr::select(
      tidyselect::all_of(id_col),
      tidyselect::all_of(pfas_names),
      tidyselect::all_of(protein_names)
    ) %>%
    dplyr::rename(participant_id = tidyselect::all_of(id_col)) %>%
    rename_pfas_to_common_names(pfas_names) %>%
    dplyr::filter(
      stats::complete.cases(dplyr::select(., tidyselect::all_of(protein_names)))
    )
}

make_scaled_matrix <- function(df, feature_names, center = NULL, scale = NULL) {
  X <- as.matrix(df[, feature_names, drop = FALSE])
  storage.mode(X) <- "double"

  if (is.null(center)) {
    center <- colMeans(X, na.rm = TRUE)
  }
  if (is.null(scale)) {
    scale <- apply(X, 2, stats::sd, na.rm = TRUE)
  }

  scale[is.na(scale) | scale == 0] <- 1

  list(
    X = sweep(sweep(X, 2, center, "-"), 2, scale, "/"),
    center = center,
    scale = scale
  )
}
```

## Complete-case model matrices

```{r praxis-data}
praxis_feature_names <- selected_proteins

SOLAR_praxis_df <- make_praxis_df(
  df = SOLAR_working_df,
  id_col = "id",
  pfas_names = SOLAR_pfas_names,
  protein_names = selected_proteins
)

CHS_praxis_df <- make_praxis_df(
  df = CHS_working_df,
  id_col = "id",
  pfas_names = CHS_pfas_names,
  protein_names = selected_proteins
)

BHRC_praxis_df <- make_praxis_df(
  df = BHRC_working_df,
  id_col = "LABID",
  pfas_names = BHRC_pfas_names,
  protein_names = selected_proteins
)

SOLAR_scaled <- make_scaled_matrix(SOLAR_praxis_df, praxis_feature_names)
CHS_scaled <- make_scaled_matrix(
  CHS_praxis_df,
  praxis_feature_names,
  center = SOLAR_scaled$center,
  scale = SOLAR_scaled$scale
)
BHRC_scaled <- make_scaled_matrix(
  BHRC_praxis_df,
  praxis_feature_names,
  center = SOLAR_scaled$center,
  scale = SOLAR_scaled$scale
)

tibble::tibble(
  cohort = cohort_names,
  n_participants = c(nrow(SOLAR_scaled$X), nrow(CHS_scaled$X), nrow(BHRC_scaled$X)),
  n_pfas_in_model = 0L,
  n_proteins = length(selected_proteins),
  n_total_features = length(praxis_feature_names)
) %>%
  kableExtra::kbl(caption = "Praxis-BGM complete-case modeling matrices.") %>%
  kableExtra::kable_paper("hover", full_width = FALSE)
```

## Input feature inventory

```{r praxis-input-features}
praxis_input_features <- tibble::tibble(
  feature_name = praxis_feature_names,
  feature_type = "Protein"
) %>%
  dplyr::left_join(proteomics_property, by = c("feature_name" = "Assay")) %>%
  dplyr::mutate(
    input_source = "PWAS/qgcomp-screened protein"
  )

praxis_input_features %>%
  dplyr::count(feature_type, input_source, name = "n_features") %>%
  kableExtra::kbl(caption = "Features passed into Praxis-BGM.") %>%
  kableExtra::kable_paper("hover", full_width = FALSE)

praxis_input_features %>%
  dplyr::filter(feature_type == "Protein") %>%
  dplyr::select(feature_name, Protein, UniProt, input_source) %>%
  dplyr::slice_head(n = 20) %>%
  kableExtra::kbl(caption = "First 20 proteins passed into Praxis-BGM.") %>%
  kableExtra::kable_paper("hover", full_width = FALSE)
```

# SOLAR Prior Model

The SOLAR model is the reference Praxis-BGM fit.
It learns two clusters from the SOLAR selected-protein matrix and produces posterior cluster parameters.
These posterior means, covariances, and weights are saved as priors for CHS and BHRC.

## Fit SOLAR reference clusters

```{r fit-solar-praxis}
set.seed(praxis_seed)

SOLAR_praxis_fit <- praxis_bgm_fit(
  data = SOLAR_scaled$X,
  K = praxis_k,
  seed = praxis_seed,
  beta = praxis_beta,
  tol = praxis_tol,
  max_iters = praxis_max_iters,
  prior_mus_variance = praxis_prior_mus_variance,
  num_samples = praxis_num_samples,
  data_precision_int = praxis_data_precision_int,
  likelihood_temp = praxis_likelihood_temp,
  rho_prec = praxis_rho_prec,
  rho_mu = praxis_rho_mu,
  elbo_eval_freq = praxis_elbo_eval_freq,
  num_iters = praxis_num_iters,
  batch_size = min(praxis_batch_size_max, nrow(SOLAR_scaled$X)),
  early_stop = praxis_early_stop,
  patience = praxis_patience,
  verbose = TRUE
)

SOLAR_cluster_assignments <- tibble::tibble(
  cohort = "SOLAR",
  participant_id = SOLAR_praxis_df$participant_id,
  praxis_cluster = paste0("C", SOLAR_praxis_fit$assignments),
  praxis_cluster_numeric = SOLAR_praxis_fit$assignments
)

SOLAR_priors <- list(
  prior_mus = SOLAR_praxis_fit$posterior_mus,
  prior_Sigmas = SOLAR_praxis_fit$posterior_covs * praxis_covariance_transfer_scale,
  prior_weights = SOLAR_praxis_fit$learned_weights
)

saveRDS(
  list(
    fit = SOLAR_praxis_fit,
    priors = SOLAR_priors,
    feature_names = praxis_feature_names,
    scaling = list(center = SOLAR_scaled$center, scale = SOLAR_scaled$scale),
    selected_proteins = selected_proteins
  ),
  fs::path(dir_praxis, "SOLAR_praxis_BGM_reference_model.rds")
)

SOLAR_cluster_assignments %>%
  dplyr::count(praxis_cluster, name = "n") %>%
  dplyr::mutate(pct = n / sum(n)) %>%
  kableExtra::kbl(caption = "SOLAR Praxis-BGM reference cluster sizes.") %>%
  kableExtra::kable_paper("hover", full_width = FALSE)
```

# CHS and BHRC Transfer Models

The CHS and BHRC models are fit in the same standardized feature space as SOLAR.
The SOLAR posterior means, covariances, and mixture weights are supplied as prior and initialization values.
This makes the target-cohort clusters interpretable relative to the SOLAR reference clusters rather than as completely separate unsupervised solutions.

## Transfer-model helper

```{r transfer-model-helper}
fit_transfer_praxis <- function(X, target_df, cohort, seed_offset) {
  fit <- praxis_bgm_fit(
    data = X,
    K = praxis_k,
    seed = praxis_seed + seed_offset,
    prior_mus = SOLAR_priors$prior_mus,
    prior_Sigmas = SOLAR_priors$prior_Sigmas,
    prior_weights = SOLAR_priors$prior_weights,
    init_mus = SOLAR_priors$prior_mus,
    init_covs = SOLAR_priors$prior_Sigmas,
    init_pis = SOLAR_priors$prior_weights,
    beta = praxis_beta,
    tol = praxis_tol,
    max_iters = praxis_max_iters,
    prior_mus_variance = praxis_prior_mus_variance,
    num_samples = praxis_num_samples,
    data_precision_int = praxis_data_precision_int,
    likelihood_temp = praxis_likelihood_temp,
    rho_prec = praxis_rho_prec,
    rho_mu = praxis_rho_mu,
    elbo_eval_freq = praxis_elbo_eval_freq,
    num_iters = praxis_num_iters,
    batch_size = min(praxis_batch_size_max, nrow(X)),
    early_stop = praxis_early_stop,
    patience = praxis_patience,
    verbose = TRUE
  )

  assignments <- tibble::tibble(
    cohort = cohort,
    participant_id = target_df$participant_id,
    praxis_cluster = paste0("C", fit$assignments),
    praxis_cluster_numeric = fit$assignments
  )

  list(fit = fit, assignments = assignments)
}
```

## Fit CHS and BHRC transfer clusters

```{r fit-transfer-praxis}
CHS_praxis_transfer <- fit_transfer_praxis(
  X = CHS_scaled$X,
  target_df = CHS_praxis_df,
  cohort = "CHS",
  seed_offset = 1
)

BHRC_praxis_transfer <- fit_transfer_praxis(
  X = BHRC_scaled$X,
  target_df = BHRC_praxis_df,
  cohort = "BHRC",
  seed_offset = 2
)

praxis_cluster_assignments <- dplyr::bind_rows(
  SOLAR_cluster_assignments,
  CHS_praxis_transfer$assignments,
  BHRC_praxis_transfer$assignments
)

readr::write_csv(
  praxis_cluster_assignments,
  fs::path(dir_praxis, "Praxis_BGM_SOLAR_prior_cluster_assignments.csv")
)

saveRDS(
  list(
    CHS = CHS_praxis_transfer,
    BHRC = BHRC_praxis_transfer,
    feature_names = praxis_feature_names,
    SOLAR_priors = SOLAR_priors,
    scaling = list(center = SOLAR_scaled$center, scale = SOLAR_scaled$scale)
  ),
  fs::path(dir_praxis, "CHS_BHRC_praxis_BGM_transfer_models.rds")
)

praxis_cluster_assignments %>%
  dplyr::count(cohort, praxis_cluster, name = "n") %>%
  dplyr::group_by(cohort) %>%
  dplyr::mutate(pct = n / sum(n)) %>%
  dplyr::ungroup() %>%
  kableExtra::kbl(caption = "Praxis-BGM cluster sizes by cohort.") %>%
  kableExtra::kable_paper("hover", full_width = FALSE)
```

# Feature Contributions

Praxis-BGM provides a Bayes-factor-based feature selection helper through the R wrapper.
This is the model-specific feature importance output: features ranked here are the selected proteins that the fitted mixture model considers most useful for separating clusters.
Because wrapper outputs can vary by package version, the table below records the feature ranking and any available Bayes-factor/statistic fields returned by the helper.

## Bayes-factor feature ranking

```{r bayes-factor-selection}
SOLAR_bf <- praxis_bgm_bf_selection(
  model = SOLAR_praxis_fit,
  data = SOLAR_scaled$X,
  top_n = length(praxis_feature_names),
  visual = FALSE
)

CHS_bf <- praxis_bgm_bf_selection(
  model = CHS_praxis_transfer$fit,
  data = CHS_scaled$X,
  top_n = length(praxis_feature_names),
  visual = FALSE
)

BHRC_bf <- praxis_bgm_bf_selection(
  model = BHRC_praxis_transfer$fit,
  data = BHRC_scaled$X,
  top_n = length(praxis_feature_names),
  visual = FALSE
)

make_bf_table <- function(bf, cohort) {
  out <- tibble::tibble(
    cohort = cohort,
    feature_index = bf$top_features,
    feature_name = praxis_feature_names[bf$top_features],
    importance_value = NA_real_
  ) %>%
    dplyr::mutate(
      feature_type = "Protein"
    )

  candidate_value_names <- c("bayes_factors", "bf_values", "BF", "scores", "feature_scores")
  value_name <- intersect(candidate_value_names, names(bf))[1]

  if (!is.na(value_name)) {
    values <- bf[[value_name]]
    if (length(values) == length(praxis_feature_names)) {
      out <- out %>%
        dplyr::mutate(importance_value = as.numeric(values[feature_index]))
    } else if (length(values) == length(bf$top_features)) {
      out <- out %>%
        dplyr::mutate(importance_value = as.numeric(values))
    }
  }

  out %>%
    dplyr::mutate(importance_rank = dplyr::row_number()) %>%
    dplyr::left_join(proteomics_property, by = c("feature_name" = "Assay"))
}

praxis_bf_top_features <- dplyr::bind_rows(
  make_bf_table(SOLAR_bf, "SOLAR"),
  make_bf_table(CHS_bf, "CHS"),
  make_bf_table(BHRC_bf, "BHRC")
)

readr::write_csv(
  praxis_bf_top_features,
  fs::path(dir_praxis, "Praxis_BGM_bayes_factor_top_features.csv")
)

praxis_bf_top_features %>%
  dplyr::select(cohort, importance_rank, tidyselect::any_of("importance_value"), feature_index, feature_type, feature_name, Protein, UniProt) %>%
  kableExtra::kbl(caption = "Top Praxis-BGM Bayes-factor features by cohort.") %>%
  kableExtra::kable_paper("hover", full_width = FALSE)
```

# Cluster Summaries

Cluster summaries describe how selected protein features differ between the two participant clusters in each cohort.
The protein table keeps the original log2 scale for interpretability.
The PFAS concentration plot is a post hoc characterization analysis: it exponentiates the log2 PFAS values back to the original concentration scale, but PFAS compounds were not used to fit Praxis-BGM.
Protein ranking visualizations use the Praxis-BGM Bayes-factor feature importance output, so the displayed ranks reflect model-based feature contribution rather than post hoc abundance contrasts.

The visualization strategy focuses on UMAP projections of protein expression, PFAS concentration profiles by cluster, Bayes-factor marker protein plots, and a rank-flow plot that shows which proteins repeatedly receive high model-based importance across cohorts.

## Summary helper

```{r cluster-summary-helper}
summarise_cluster_features <- function(df, assignments, feature_names, cohort) {
  df %>%
    dplyr::select(participant_id, tidyselect::all_of(feature_names)) %>%
    dplyr::left_join(assignments, by = "participant_id") %>%
    tidyr::pivot_longer(
      cols = tidyselect::all_of(feature_names),
      names_to = "feature_name",
      values_to = "value"
    ) %>%
    dplyr::group_by(cohort, praxis_cluster, feature_name) %>%
    dplyr::summarise(
      n = sum(!is.na(value)),
      mean = mean(value, na.rm = TRUE),
      sd = stats::sd(value, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    dplyr::mutate(feature_type = "Protein")
}
```

## Combined plotting data

```{r plot-data}
make_feature_df <- function(df, assignments, cohort) {
  df %>%
    dplyr::select(participant_id, tidyselect::all_of(praxis_feature_names)) %>%
    dplyr::left_join(assignments, by = "participant_id") %>%
    dplyr::mutate(cohort = cohort)
}

praxis_feature_df <- dplyr::bind_rows(
  make_feature_df(SOLAR_praxis_df, SOLAR_cluster_assignments, "SOLAR"),
  make_feature_df(CHS_praxis_df, CHS_praxis_transfer$assignments, "CHS"),
  make_feature_df(BHRC_praxis_df, BHRC_praxis_transfer$assignments, "BHRC")
)

make_pfas_df <- function(df, assignments, cohort) {
  df %>%
    dplyr::select(participant_id, tidyselect::all_of(PFAS_abbrev_names)) %>%
    dplyr::left_join(assignments, by = "participant_id") %>%
    dplyr::mutate(cohort = cohort)
}

praxis_pfas_df <- dplyr::bind_rows(
  make_pfas_df(SOLAR_praxis_df, SOLAR_cluster_assignments, "SOLAR"),
  make_pfas_df(CHS_praxis_df, CHS_praxis_transfer$assignments, "CHS"),
  make_pfas_df(BHRC_praxis_df, BHRC_praxis_transfer$assignments, "BHRC")
)
```

## Protein Cluster Summary Table

```{r cluster-summaries}
praxis_cluster_feature_summary <- dplyr::bind_rows(
  summarise_cluster_features(SOLAR_praxis_df, SOLAR_cluster_assignments, praxis_feature_names, "SOLAR"),
  summarise_cluster_features(CHS_praxis_df, CHS_praxis_transfer$assignments, praxis_feature_names, "CHS"),
  summarise_cluster_features(BHRC_praxis_df, BHRC_praxis_transfer$assignments, praxis_feature_names, "BHRC")
)

praxis_cluster_feature_summary %>%
  dplyr::filter(feature_type == "Protein") %>%
  dplyr::select(cohort, praxis_cluster, feature_name, n, mean, sd) %>%
  dplyr::slice_head(n = 20) %>%
  kableExtra::kbl(caption = "First 20 selected-protein summaries by Praxis-BGM cluster.") %>%
  kableExtra::kable_paper("hover", full_width = FALSE)
```

## Plot helper

```{r plot-helper}
praxis_cluster_colors <- c(
  C1 = "#E69F00",
  C2 = "#009E73"
)

praxis_plot_theme <- function(...) {
  theme_args <- utils::modifyList(
    list(
      plot.title = ggplot2::element_text(size = 10),
      axis.title = ggplot2::element_text(size = 10),
      axis.text = ggplot2::element_text(size = 9),
      axis.text.y = ggplot2::element_text(size = 8),
      strip.text = ggplot2::element_text(size = 9),
      legend.title = ggplot2::element_text(size = 9),
      legend.text = ggplot2::element_text(size = 9),
      legend.position = "bottom",
      panel.grid.minor = ggplot2::element_blank()
    ),
    list(...)
  )

  ggplot2::theme_bw(base_size = 9) +
    do.call(ggplot2::theme, theme_args)
}
```

## Cluster size distribution

```{r cluster-size-plot, fig.width=8, fig.height=4}
praxis_cluster_assignments %>%
  dplyr::count(cohort, praxis_cluster, name = "n") %>%
  dplyr::group_by(cohort) %>%
  dplyr::mutate(pct = n / sum(n)) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    cohort = factor(cohort, levels = c("SOLAR", "CHS", "BHRC")),
    praxis_cluster = factor(praxis_cluster, levels = names(praxis_cluster_colors))
  ) %>%
  ggplot2::ggplot(ggplot2::aes(x = cohort, y = pct, fill = praxis_cluster)) +
  ggplot2::geom_col(width = 0.65, color = "white") +
  ggplot2::geom_text(
    ggplot2::aes(label = scales::percent(pct, accuracy = 1)),
    position = ggplot2::position_stack(vjust = 0.5),
    color = "white",
    size = 3.1
  ) +
  ggplot2::scale_fill_manual(values = praxis_cluster_colors) +
  ggplot2::scale_y_continuous(labels = scales::percent_format()) +
  ggplot2::labs(
    x = NULL,
    y = "Participants",
    fill = "Praxis-BGM cluster"
  ) +
  praxis_plot_theme(
    legend.box.spacing = grid::unit(0, "pt"),
    legend.margin = ggplot2::margin(0, 0, 0, 0),
    legend.spacing.y = grid::unit(0, "pt"),
    legend.key.height = grid::unit(8, "pt"),
    plot.margin = ggplot2::margin(5.5, 5.5, 0, 5.5)
  )
```

```{r}
#| eval: false

ggsave(
  filename = fs::path(
    dir_figures,
    "praxis_BGM_cluster_distribution.png"
  ),
  width = 3, height = 2.5, dpi = 300, units = "in"
)
```

## Protein UMAP

```{r protein-umap, fig.width=10, fig.height=7}
protein_umap_available <- requireNamespace("uwot", quietly = TRUE) || requireNamespace("umap", quietly = TRUE)

if (!protein_umap_available) {
  message("Skipping UMAP because neither the 'uwot' nor 'umap' package is installed.")
} else {
  pooled_protein_matrix <- rbind(
    SOLAR_scaled$X[, selected_proteins, drop = FALSE],
    CHS_scaled$X[, selected_proteins, drop = FALSE],
    BHRC_scaled$X[, selected_proteins, drop = FALSE]
  )

  pooled_protein_meta <- dplyr::bind_rows(
    SOLAR_cluster_assignments,
    CHS_praxis_transfer$assignments,
    BHRC_praxis_transfer$assignments
  )

  set.seed(praxis_seed)

  if (requireNamespace("uwot", quietly = TRUE)) {
    pooled_umap <- uwot::umap(
      pooled_protein_matrix,
      n_neighbors = min(15, nrow(pooled_protein_matrix) - 1),
      min_dist = 0.1,
      metric = "euclidean",
      ret_model = FALSE,
      verbose = FALSE
    )
  } else {
    pooled_umap <- umap::umap(pooled_protein_matrix)$layout
  }

  praxis_umap_df <- tibble::tibble(
    UMAP1 = pooled_umap[, 1],
    UMAP2 = pooled_umap[, 2]
  ) %>%
    dplyr::bind_cols(pooled_protein_meta) %>%
    dplyr::mutate(
      cohort = factor(cohort, levels = c("SOLAR", "CHS", "BHRC")),
      praxis_cluster = factor(praxis_cluster, levels = names(praxis_cluster_colors))
    )

  praxis_umap_df %>%
    ggplot2::ggplot(ggplot2::aes(x = UMAP1, y = UMAP2, color = praxis_cluster, shape = cohort)) +
    ggplot2::geom_point(alpha = 0.8, size = 1, stroke = 0.8) +
    ggplot2::scale_color_manual(values = praxis_cluster_colors) +
    ggplot2::scale_shape_manual(values = c(SOLAR = 16, CHS = 17, BHRC = 4)) +
    ggplot2::labs(
      x = "UMAP 1",
      y = "UMAP 2",
      color = "Praxis-BGM cluster",
      shape = "Cohort"
    ) +
    ggplot2::guides(
      color = ggplot2::guide_legend(nrow = 1, order = 1),
      shape = ggplot2::guide_legend(nrow = 1, order = 2)
    ) +
    praxis_plot_theme(
      legend.box = "vertical",
      legend.box.spacing = grid::unit(1, "pt"),
      legend.spacing.y = grid::unit(1, "pt"),
      legend.margin = ggplot2::margin(0, 0, 0, 0),
      legend.key.height = grid::unit(8, "pt")
    )
}
```

```{r}
#| eval: false

ggsave(
  filename = fs::path(
    dir_figures,
    "praxis_BGM_UMAP.png"
  ),
  width = 3, height = 2.5, dpi = 300, units = "in"
)
```

## PFAS concentration profiles

```{r cluster-pfas-plot, fig.width=10, fig.height=6}
pfas_cluster_values <- praxis_pfas_df %>%
  dplyr::select(cohort, participant_id, praxis_cluster, tidyselect::all_of(PFAS_abbrev_names)) %>%
  tidyr::pivot_longer(
    cols = tidyselect::all_of(PFAS_abbrev_names),
    names_to = "feature_name",
    values_to = "log2_value"
  ) %>%
  dplyr::mutate(
    concentration = 2^log2_value,
    cohort = factor(cohort, levels = c("SOLAR", "CHS", "BHRC")),
    praxis_cluster = factor(praxis_cluster, levels = c("C1", "C2")),
    feature_name = factor(feature_name, levels = PFAS_abbrev_names)
  )

pfas_cluster_summary <- pfas_cluster_values %>%
  dplyr::group_by(cohort, feature_name, praxis_cluster) %>%
  dplyr::summarise(
    mean_concentration = mean(concentration, na.rm = TRUE),
    se_concentration = stats::sd(concentration, na.rm = TRUE) / sqrt(sum(!is.na(concentration))),
    .groups = "drop"
  )

pfas_cluster_tests <- pfas_cluster_values %>%
  dplyr::filter(!is.na(concentration), !is.na(praxis_cluster)) %>%
  dplyr::group_by(cohort, feature_name) %>%
  dplyr::group_modify(function(.x, .y) {
    p_value <- if (dplyr::n_distinct(.x$praxis_cluster) == 2 && all(table(.x$praxis_cluster) >= 2)) {
      tryCatch(
        stats::t.test(concentration ~ praxis_cluster, data = .x)$p.value,
        error = function(e) NA_real_
      )
    } else {
      NA_real_
    }

    tibble::tibble(p_value = p_value)
  }) %>%
  dplyr::ungroup() %>%
  dplyr::group_by(cohort) %>%
  dplyr::mutate(p_adjust_BH = stats::p.adjust(p_value, method = "BH")) %>%
  dplyr::ungroup()

pfas_cluster_annotation <- pfas_cluster_summary %>%
  dplyr::group_by(cohort, feature_name) %>%
  dplyr::slice_max(mean_concentration, n = 1, with_ties = FALSE) %>%
  dplyr::ungroup() %>%
  dplyr::left_join(pfas_cluster_tests, by = c("cohort", "feature_name")) %>%
  dplyr::filter(!is.na(p_value), p_value < 0.05) %>%
  dplyr::mutate(label_y = mean_concentration + se_concentration)

pfas_cluster_summary %>%
  ggplot2::ggplot(ggplot2::aes(x = feature_name, y = mean_concentration)) +
  ggplot2::geom_col(
    ggplot2::aes(fill = praxis_cluster),
    position = ggplot2::position_dodge(width = 0.72),
    width = 0.62
  ) +
  ggplot2::geom_errorbar(
    ggplot2::aes(
      ymin = pmax(mean_concentration - se_concentration, 0),
      ymax = mean_concentration + se_concentration,
      color = praxis_cluster
    ),
    position = ggplot2::position_dodge(width = 0.72),
    width = 0.18,
    linewidth = 0.35
  ) +
  ggplot2::geom_text(
    data = pfas_cluster_annotation,
    ggplot2::aes(x = feature_name, y = label_y, label = "*", group = praxis_cluster),
    inherit.aes = FALSE,
    position = ggplot2::position_dodge(width = 0.72),
    vjust = -0.2,
    size = 3.1
  ) +
  ggplot2::facet_grid(. ~ cohort) +
  ggplot2::scale_fill_manual(values = praxis_cluster_colors) +
  ggplot2::scale_color_manual(values = praxis_cluster_colors) +
  ggplot2::scale_y_sqrt(
    labels = scales::label_number(accuracy = 0.01),
    expand = ggplot2::expansion(mult = c(0, 0.12))
  ) +
  ggplot2::labs(
    x = NULL,
    y = paste0("PFAS concentration (", pfas_concentration_unit, ")\nsquare-root scale"),
    fill = "Praxis-BGM cluster",
    color = "Praxis-BGM cluster"
  ) +
  ggplot2::guides(color = "none") +
  praxis_plot_theme(
    legend.box.spacing = grid::unit(1, "pt"),
    legend.margin = ggplot2::margin(0, 0, 0, 0),
    legend.spacing.x = grid::unit(2, "pt"),
    legend.spacing.y = grid::unit(1, "pt"),
    legend.key.height = grid::unit(8, "pt"),
    axis.text.x = ggplot2::element_text(size = 9, angle = 90, vjust = 0.5, hjust = 1)
  )
```

```{r}
#| eval: false

ggsave(
  filename = fs::path(
    dir_figures,
    "praxis_BGM_pfas_profiles.png"
  ),
  width = 4, height = 2.5, units = "in"
)
```

## Protein Bayes-factor ranks

```{r cluster-contrast-importance}
calculate_cluster_contrasts <- function(df, assignments, cohort) {
  feature_df <- df %>%
    dplyr::select(participant_id, tidyselect::all_of(praxis_feature_names)) %>%
    dplyr::left_join(assignments, by = "participant_id") %>%
    dplyr::filter(!is.na(praxis_cluster))

  cluster_levels <- sort(unique(feature_df$praxis_cluster))

  if (length(cluster_levels) != 2) {
    return(tibble::tibble())
  }

  means <- feature_df %>%
    dplyr::group_by(praxis_cluster) %>%
    dplyr::summarise(dplyr::across(tidyselect::all_of(praxis_feature_names), ~ mean(.x, na.rm = TRUE)), .groups = "drop") %>%
    tidyr::pivot_longer(
      cols = tidyselect::all_of(praxis_feature_names),
      names_to = "feature_name",
      values_to = "mean_value"
    )

  means %>%
    tidyr::pivot_wider(names_from = praxis_cluster, values_from = mean_value) %>%
    dplyr::mutate(
      cohort = cohort,
      contrast = .data[[cluster_levels[2]]] - .data[[cluster_levels[1]]],
      abs_contrast = abs(contrast),
      feature_type = "Protein"
    ) %>%
    dplyr::left_join(proteomics_property, by = c("feature_name" = "Assay")) %>%
    dplyr::arrange(dplyr::desc(abs_contrast))
}

praxis_cluster_contrasts <- dplyr::bind_rows(
  calculate_cluster_contrasts(SOLAR_praxis_df, SOLAR_cluster_assignments, "SOLAR"),
  calculate_cluster_contrasts(CHS_praxis_df, CHS_praxis_transfer$assignments, "CHS"),
  calculate_cluster_contrasts(BHRC_praxis_df, BHRC_praxis_transfer$assignments, "BHRC")
)

praxis_protein_ranks <- praxis_bf_top_features %>%
  dplyr::filter(feature_type == "Protein") %>%
  dplyr::group_by(cohort) %>%
  dplyr::arrange(importance_rank, .by_group = TRUE) %>%
  dplyr::mutate(cohort_rank = dplyr::row_number()) %>%
  dplyr::ungroup() %>%
  dplyr::left_join(
    praxis_cluster_contrasts %>%
      dplyr::select(cohort, feature_name, contrast, abs_contrast),
    by = c("cohort", "feature_name")
  )

praxis_protein_ranks %>%
  dplyr::group_by(cohort) %>%
  dplyr::slice_min(cohort_rank, n = 20, with_ties = FALSE) %>%
  dplyr::ungroup() %>%
  dplyr::select(cohort, cohort_rank, importance_value, feature_name, Protein, UniProt, contrast, abs_contrast) %>%
  kableExtra::kbl(caption = "Top proteins by Praxis-BGM Bayes-factor feature importance.") %>%
  kableExtra::kable_paper("hover", full_width = FALSE)
```

## Robust Rank Aggregation

Robust Rank Aggregation tests whether proteins repeatedly appear near the top of the cohort-specific Praxis-BGM Bayes-factor rankings more often than expected by chance.
Here, each cohort contributes one complete ranked list of all protein features used in Praxis-BGM, ordered from highest to lowest Bayes-factor importance.
Lower RRA p-values indicate proteins with more consistent high ranks across SOLAR, CHS, and BHRC.

```{r robust-rank-aggregation}
if (!requireNamespace("RobustRankAggreg", quietly = TRUE)) {
  stop(
    "The 'RobustRankAggreg' package is required for Robust Rank Aggregation. ",
    "Install it with install.packages('RobustRankAggreg')."
  )
}

rra_ranked_lists <- praxis_protein_ranks %>%
  dplyr::filter(!is.na(cohort_rank)) %>%
  dplyr::mutate(cohort = factor(cohort, levels = c("SOLAR", "CHS", "BHRC"))) %>%
  dplyr::arrange(cohort, cohort_rank) %>%
  dplyr::group_by(cohort) %>%
  dplyr::summarise(ranked_features = list(feature_name), .groups = "drop")

rra_feature_lists <- rra_ranked_lists$ranked_features
names(rra_feature_lists) <- as.character(rra_ranked_lists$cohort)

rra_raw <- RobustRankAggreg::aggregateRanks(
  glist = rra_feature_lists,
  N = length(praxis_feature_names),
  method = "RRA"
)

rra_results <- tibble::as_tibble(rra_raw)
if (!"Name" %in% names(rra_results)) {
  rra_results <- rra_raw %>%
    as.data.frame() %>%
    tibble::rownames_to_column("Name") %>%
    tibble::as_tibble()
}

praxis_rra_results <- rra_results %>%
  dplyr::rename(
    feature_name = Name,
    rra_p_value = Score
  ) %>%
  dplyr::mutate(rra_adjusted_pval_BH = stats::p.adjust(rra_p_value, method = "BH")) %>%
  dplyr::left_join(proteomics_property, by = c("feature_name" = "Assay")) %>%
  dplyr::left_join(
    praxis_protein_ranks %>%
      dplyr::select(cohort, feature_name, cohort_rank, importance_value) %>%
      tidyr::pivot_wider(
        names_from = cohort,
        values_from = c(cohort_rank, importance_value),
        names_glue = "{cohort}_{.value}"
      ),
    by = "feature_name"
  ) %>%
  dplyr::arrange(rra_p_value)

praxis_rra_results %>%
  dplyr::transmute(
    `Olink assay` = feature_name,
    `Protein name` = Protein,
    `UniProt ID` = UniProt,
    `RRA p-value` = scales::pvalue(rra_p_value, accuracy = 0.001),
    `BH-adjusted RRA p-value` = scales::pvalue(rra_adjusted_pval_BH, accuracy = 0.001),
    `SOLAR rank` = SOLAR_cohort_rank,
    `CHS rank` = CHS_cohort_rank,
    `BHRC rank` = BHRC_cohort_rank,
    `SOLAR importance` = round(SOLAR_importance_value, 2),
    `CHS importance` = round(CHS_importance_value, 2),
    `BHRC importance` = round(BHRC_importance_value, 2)
  ) %>%
  dplyr::slice_head(n = 20) %>%
  kableExtra::kbl(
    caption = "Supplemental Table. Top proteins from Robust Rank Aggregation of Praxis-BGM protein importance rankings across SOLAR, CHS, and BHRC.",
    align = c("l", "l", "l", rep("r", 8))
  ) %>%
  kableExtra::kable_paper("hover", full_width = FALSE) %>%
  kableExtra::footnote(
    general = paste(
      "RRA, Robust Rank Aggregation; BH, Benjamini-Hochberg.",
      "Praxis-BGM Bayes factors were used as model-based protein importance scores, with larger values indicating greater contribution to cluster discrimination.",
      "Each cohort-specific rank is based on Praxis-BGM Bayes-factor feature importance, with rank 1 indicating the highest-ranked protein within that cohort.",
      "Lower RRA p-values indicate proteins that appear near the top of cohort-specific rankings more consistently than expected by chance."
    ),
    general_title = "Note: ",
    footnote_as_chunk = TRUE
  )
```

## Marker protein Reactome annotation

```{r marker-protein-reactome-annotation}
marker_selected_features <- praxis_protein_ranks %>%
  dplyr::filter(!is.na(importance_value), importance_value >= marker_bf_threshold) %>%
  dplyr::mutate(
    cohort = factor(cohort, levels = c("SOLAR", "CHS", "BHRC")),
    display_name = dplyr::coalesce(Protein, feature_name)
  ) %>%
  dplyr::arrange(cohort, dplyr::desc(importance_value), cohort_rank) %>%
  dplyr::mutate(display_name = factor(display_name, levels = unique(display_name))) %>%
  dplyr::select(cohort, feature_name, display_name, cohort_rank, importance_value)

if (nrow(marker_selected_features) > 0) {
  marker_reactome_input <- marker_selected_features %>%
    dplyr::transmute(
      protein = feature_name,
      protein_label = as.character(display_name)
    ) %>%
    dplyr::distinct()

  marker_reactome_assignment <- annotate_reactome_pathways(
    marker_reactome_input,
    n_pathways_to_show = c(5, 10)
  )

  marker_reactome_pathways <- marker_reactome_assignment$pathways %>%
    dplyr::distinct(protein, reactome_pathway_id, reactome_pathway) %>%
    dplyr::mutate(
      reactome_pathway = factor(reactome_pathway, levels = sort(unique(reactome_pathway)))
    )

  marker_reactome_pathway_colors <- make_reactome_pathway_colors(
    levels(marker_reactome_pathways$reactome_pathway)
  )

  recurring_marker_labels <- marker_selected_features %>%
    dplyr::count(feature_name, display_name) %>%
    dplyr::filter(n > 1) %>%
    dplyr::pull(display_name) %>%
    as.character()

  marker_reactome_pathways %>%
    dplyr::count(reactome_pathway, name = "n_marker_proteins") %>%
    dplyr::arrange(dplyr::desc(n_marker_proteins), reactome_pathway) %>%
    kableExtra::kbl(caption = "Reactome pathway assignments for marker proteins.") %>%
    kableExtra::kable_paper("hover", full_width = FALSE)
} else {
  marker_reactome_pathways <- tibble::tibble(
    protein = character(),
    reactome_pathway_id = character(),
    reactome_pathway = factor(character())
  )
  marker_reactome_pathway_colors <- character()
  recurring_marker_labels <- character()
}
```

## Marker protein Bayes-factor plot

```{r marker-protein-bayes-factor-plot, fig.width=10, fig.height=8}
if (nrow(marker_selected_features) == 0) {
  message(
    "Skipping marker protein abundance plot: no proteins have Praxis-BGM ",
    "Bayes-factor importance >= ", marker_bf_threshold, "."
  )
} else {
  if (!requireNamespace("ggtext", quietly = TRUE)) {
    stop(
      "The 'ggtext' package is required for Reactome-colored y-axis labels. ",
      "Install it with install.packages('ggtext')."
    )
  }

  marker_abundance_df <- praxis_feature_df %>%
    dplyr::select(cohort, participant_id, praxis_cluster, tidyselect::all_of(unique(marker_selected_features$feature_name))) %>%
    tidyr::pivot_longer(
      cols = tidyselect::all_of(unique(marker_selected_features$feature_name)),
      names_to = "feature_name",
      values_to = "protein_abundance"
    ) %>%
    dplyr::mutate(cohort = factor(cohort, levels = c("SOLAR", "CHS", "BHRC"))) %>%
    dplyr::inner_join(
      marker_selected_features %>%
        dplyr::left_join(
          marker_reactome_pathways,
          by = c("feature_name" = "protein")
        ),
      by = c("cohort", "feature_name")
    )

  escape_markdown_label <- function(x) {
    x %>%
      stringr::str_replace_all("&", "&amp;") %>%
      stringr::str_replace_all("<", "&lt;") %>%
      stringr::str_replace_all(">", "&gt;")
  }

  marker_pathway_lookup <- marker_abundance_df %>%
    dplyr::distinct(cohort, display_name, reactome_pathway) %>%
    dplyr::mutate(
      display_name = as.character(display_name),
      legend_x = min(marker_abundance_df$protein_abundance, na.rm = TRUE)
    )

  format_marker_axis_labels <- function(x) {
    label_text <- escape_markdown_label(x)
    label_pathway <- marker_pathway_lookup$reactome_pathway[
      match(x, marker_pathway_lookup$display_name)
    ]
    label_color <- marker_reactome_pathway_colors[as.character(label_pathway)]
    label_color[is.na(label_color)] <- "grey30"
    label_style <- dplyr::if_else(
      x %in% recurring_marker_labels,
      "font-style:italic;",
      ""
    )

    paste0(
      "<span style='color:", label_color, ";", label_style, "'>",
      label_text,
      "</span>"
    )
  }

  marker_abundance_df %>%
    ggplot2::ggplot(ggplot2::aes(x = protein_abundance, y = display_name, fill = praxis_cluster)) +
    ggplot2::geom_boxplot(
      width = 0.65,
      outlier.size = 0.4,
      outlier.alpha = 0.45
    ) +
    ggplot2::geom_point(
      data = marker_pathway_lookup,
      ggplot2::aes(x = legend_x, y = display_name, color = reactome_pathway),
      inherit.aes = FALSE,
      alpha = 0,
      show.legend = TRUE
    ) +
    ggplot2::scale_fill_manual(values = praxis_cluster_colors) +
    ggplot2::facet_wrap(~cohort, nrow = 1, scales = "free_y") +
    ggplot2::scale_y_discrete(labels = format_marker_axis_labels) +
    ggplot2::scale_color_manual(
      values = marker_reactome_pathway_colors,
      name = "Reactome pathway"
    ) +
    ggplot2::coord_cartesian(clip = "off") +
    ggplot2::labs(
      x = "Protein abundance (log2 normalized)",
      y = NULL
    ) +
    ggplot2::guides(
      color = ggplot2::guide_legend(
        nrow = 3,
        byrow = TRUE,
        override.aes = list(alpha = 1, size = 2),
        title.position = "top",
        title.hjust = 0
      ),
      fill = "none"
    ) +
    praxis_plot_theme(legend.position = "bottom") +
    ggplot2::theme(
      axis.text.y = ggtext::element_markdown(size = 8),
      legend.box.spacing = grid::unit(1, "pt"),
      legend.margin = ggplot2::margin(0, 0, 0, 0),
      legend.spacing.x = grid::unit(2, "pt"),
      legend.key.height = grid::unit(8, "pt"),
      plot.margin = ggplot2::margin(5.5, 5.5, 5.5, 5.5)
    )
}
```

```{r}
ggsave(
  filename = fs::path(
    dir_figures,
    "praxis_BGM_marker_protein.png"
  ),
  width = 7, height = 3, dpi = 600, units = "in"
)
```

## Cross-cohort protein rank flow

```{r top-protein-rank-flow-plot, fig.width=12, fig.height=7}
rank_flow_proteins <- praxis_protein_ranks %>%
  dplyr::filter(!is.na(importance_value), importance_value >= marker_bf_threshold) %>%
  dplyr::pull(feature_name)

if (length(rank_flow_proteins) == 0) {
  message(
    "Skipping cross-cohort protein rank flow: no proteins have Praxis-BGM ",
    "Bayes-factor importance >= ", marker_bf_threshold, "."
  )
} else {
  rank_flow_df <- praxis_protein_ranks %>%
    dplyr::filter(feature_name %in% unique(rank_flow_proteins)) %>%
    dplyr::left_join(
      marker_reactome_pathways,
      by = c("feature_name" = "protein")
    ) %>%
    dplyr::mutate(
      cohort = factor(cohort, levels = c("SOLAR", "CHS", "BHRC")),
      reactome_pathway = factor(
        reactome_pathway,
        levels = names(marker_reactome_pathway_colors)
      ),
      x_position = dplyr::case_when(
        cohort == "SOLAR" ~ 0,
        cohort == "CHS" ~ 2,
        cohort == "BHRC" ~ 4
      ),
      display_name = dplyr::coalesce(Protein, feature_name),
      rank_label = paste0(cohort_rank, "  ", display_name)
    ) %>%
    dplyr::group_by(cohort) %>%
    dplyr::arrange(cohort_rank, .by_group = TRUE) %>%
    dplyr::mutate(plot_rank = dplyr::row_number()) %>%
    dplyr::ungroup()

  rank_flow_segments <- rank_flow_df %>%
    dplyr::arrange(feature_name, cohort) %>%
    dplyr::group_by(feature_name) %>%
    dplyr::mutate(
      xend_position = dplyr::lead(x_position),
      plot_rank_end = dplyr::lead(plot_rank)
    ) %>%
    dplyr::ungroup() %>%
    dplyr::filter(!is.na(xend_position), !is.na(plot_rank_end))

  rank_flow_max_rank <- max(rank_flow_df$plot_rank, na.rm = TRUE)

  ggplot2::ggplot() +
    ggplot2::geom_segment(
      data = rank_flow_segments,
      ggplot2::aes(
        x = x_position + 0.46,
        xend = xend_position - 0.46,
        y = -plot_rank,
        yend = -plot_rank_end
      ),
      color = "grey55",
      linewidth = 0.35,
      alpha = 0.75
    ) +
    ggplot2::geom_tile(
      data = rank_flow_df,
      ggplot2::aes(x = x_position, y = -plot_rank, fill = reactome_pathway),
      width = 0.82,
      height = 0.78,
      color = "grey70",
      linewidth = 0.2
    ) +
    ggplot2::geom_text(
      data = rank_flow_df,
      ggplot2::aes(x = x_position, y = -plot_rank, label = rank_label),
      color = "white",
      size = 3
    ) +
    ggplot2::annotate("text", x = 0, y = 0, label = "SOLAR", fontface = "plain", size = 3.1) +
    ggplot2::annotate("text", x = 2, y = 0, label = "CHS", fontface = "plain", size = 3.1) +
    ggplot2::annotate("text", x = 4, y = 0, label = "BHRC", fontface = "plain", size = 3.1) +
    ggplot2::annotate(
      "text",
      x = 2,
      y = -rank_flow_max_rank - 1.1,
      label = "Cross-cohort protein rank flow by Bayes-factor importance",
      size = 3.1
    ) +
    ggplot2::scale_fill_manual(
      values = marker_reactome_pathway_colors,
      name = "Reactome pathway",
      drop = FALSE
    ) +
    ggplot2::guides(
      fill = ggplot2::guide_legend(
        nrow = 3,
        byrow = TRUE,
        title.position = "top",
        title.hjust = 0
      )
    ) +
    ggplot2::coord_cartesian(
      xlim = c(-0.5, 4.5),
      ylim = c(-rank_flow_max_rank - 1.6, 0.6),
      clip = "off"
    ) +
    ggplot2::theme_void(base_size = 9) +
    ggplot2::theme(
      legend.position = "bottom",
      legend.box.spacing = grid::unit(1, "pt"),
      legend.margin = ggplot2::margin(0, 0, 0, 0),
      legend.spacing.x = grid::unit(2, "pt"),
      legend.key.height = grid::unit(8, "pt"),
      legend.title = ggplot2::element_text(size = 9),
      legend.text = ggplot2::element_text(size = 9),
      plot.margin = ggplot2::margin(0, 5, 0, 0),
      text = ggplot2::element_text(size = 9)
    )
}
```

```{r}
ggsave(
  filename = fs::path(
    dir_figures,
    "praxis_BGM_protein_rank_flow.png"
  ),
  width = 6, height = 3.5, units = "in"
)
```

## Cross-cohort protein rank bump chart

```{r protein-rank-bump-chart, fig.width=7, fig.height=4.5}
bump_chart_proteins <- praxis_protein_ranks %>%
  dplyr::filter(!is.na(importance_value), importance_value >= marker_bf_threshold) %>%
  dplyr::distinct(feature_name)

if (nrow(bump_chart_proteins) == 0) {
  message(
    "Skipping cross-cohort protein rank bump chart: no proteins have Praxis-BGM ",
    "Bayes-factor importance >= ", marker_bf_threshold, "."
  )
} else {
  bump_chart_df <- praxis_protein_ranks %>%
    dplyr::semi_join(bump_chart_proteins, by = "feature_name") %>%
    dplyr::left_join(
      marker_reactome_pathways,
      by = c("feature_name" = "protein")
    ) %>%
    dplyr::mutate(
      cohort = factor(cohort, levels = c("SOLAR", "CHS", "BHRC")),
      cohort_x = as.numeric(cohort) + 0.65,
      display_name = dplyr::coalesce(Protein, feature_name),
      reactome_pathway = factor(
        reactome_pathway,
        levels = names(marker_reactome_pathway_colors)
      )
    ) %>%
    dplyr::group_by(feature_name) %>%
    dplyr::filter(dplyr::n_distinct(cohort_rank[!is.na(cohort_rank)]) > 1) %>%
    dplyr::ungroup() %>%
    dplyr::group_by(cohort) %>%
    dplyr::arrange(cohort_rank, .by_group = TRUE) %>%
    dplyr::mutate(plot_rank = dplyr::row_number()) %>%
    dplyr::ungroup()

  if (nrow(bump_chart_df) == 0) {
    message("Skipping cross-cohort protein rank bump chart: selected protein ranks do not change across cohorts.")
  } else {
    bump_chart_labels <- bump_chart_df %>%
      dplyr::filter(cohort == "SOLAR") %>%
      dplyr::arrange(cohort_rank) %>%
      dplyr::distinct(feature_name, .keep_all = TRUE)

    protein_rank_bump_chart <- bump_chart_df %>%
      ggplot2::ggplot(
        ggplot2::aes(
          x = cohort_x,
          y = -plot_rank,
          group = feature_name,
          color = reactome_pathway
        )
      ) +
      ggplot2::geom_line(linewidth = 0.45, alpha = 0.8) +
      ggplot2::geom_point(size = 1.8, alpha = 0.95) +
      ggplot2::geom_text(
        data = bump_chart_labels,
        ggplot2::aes(x = 1.48, y = -plot_rank, label = display_name),
        inherit.aes = FALSE,
        hjust = 1,
        size = 2.6,
        color = "black",
        check_overlap = TRUE
      ) +
      ggplot2::annotate(
        "text",
        x = seq_along(cohort_names) + 0.65,
        y = 0,
        label = cohort_names,
        size = 3,
        color = "black"
      ) +
      ggplot2::scale_x_continuous(
        limits = c(0.2, length(cohort_names) + 0.85),
        expand = ggplot2::expansion(mult = 0)
      ) +
      ggplot2::scale_color_manual(
        values = marker_reactome_pathway_colors,
        name = "Reactome pathway",
        na.value = "grey60",
        drop = FALSE
      ) +
      ggplot2::coord_cartesian(
        ylim = c(-max(bump_chart_df$plot_rank, na.rm = TRUE) - 0.4, 0.4),
        clip = "off"
      ) +
      ggplot2::guides(
        color = ggplot2::guide_legend(
          nrow = 3,
          byrow = TRUE,
          title.position = "top",
          title.hjust = 0
        )
      ) +
      ggplot2::theme_void(base_size = 9) +
      ggplot2::theme(
        legend.position = "bottom",
        legend.box.spacing = grid::unit(1, "pt"),
        legend.margin = ggplot2::margin(0, 0, 0, 0),
        legend.spacing.x = grid::unit(2, "pt"),
        legend.key.height = grid::unit(8, "pt"),
        legend.title = ggplot2::element_text(size = 10, color = "black"),
        legend.text = ggplot2::element_text(size = 10, color = "black"),
        text = ggplot2::element_text(color = "black"),
        plot.margin = ggplot2::margin(0, 0, 0, 0)
      )

    protein_rank_bump_chart
  }
}
```

```{r}
if (exists("protein_rank_bump_chart")) {
  ggsave(
    plot = protein_rank_bump_chart,
    filename = fs::path(
      dir_figures,
      "praxis_BGM_protein_rank_bump_chart.png"
    ),
    width = 7, height = 3.5, dpi = 600, units = "in"
  )
}
```
