---
title: "SOLAR, CHS, and BHRC Proteomics WGCNA"
author: "Zhenjiang Li"
date: today
toc: true
toc_expand: 2
format:
  html:
    code-fold: true
    html-math-method: katex
    fontsize: 16pt
    grid:
      sidebar-width: 0px
      body-width: 1500px
      margin-width: 500px
      gutter-width: 1.5rem
embed-resources: true
editor: visual
editor_options:
  chunk_output_type: console
---

## Environment Setup {.hidden}

```{=html}
<style type="text/css">
caption, .table-caption {
  text-align: left;
}
</style>
```

```{r setup}
#| echo: false

options(scipen = 999)

library(fs)
library(here)
i_am("1_scripts/!analytic_workflow_proteomcis_WGCNA.qmd")

source(fs::path("1_scripts", "!libraries.R"))
source(fs::path("1_scripts", "!directories.R"))
source(fs::path("1_scripts", "!functions.R"))

options(knitr.table.format = "html")
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

```{r}
#| label: shorten-long-path
#| include: false

create_shortcut <- function(path, short_cut, drive) {
  return(
    sub(gsub("\\)", "\\\\)", gsub("\\(", "\\\\(", short_cut)), paste0(drive, ":"), path)
  )
}

if (nchar(dir_data) > 80) {
  system("subst x: /D")
  short_cut <- dir_home |> dirname()
  system(paste0("subst x: \"", short_cut, "\""))

  dir_objects_names <- ls(pattern = "^dir_")
  dir_list <- mget(dir_objects_names)

  dir_list_shortened <- lapply(
    dir_list,
    function(y) create_shortcut(y, short_cut, "x")
  )
  list2env(dir_list_shortened, envir = .GlobalEnv)
}

rm(list = intersect(
  c("dir_objects_names", "short_cut", "dir_list", "dir_list_shortened"),
  ls()
))
```

# Parameters

```{r}
cohort_names <- c("SOLAR", "CHS", "BHRC")
proteomics_metadata_cols <- c("LABID", "study", "Batch")

max_protein_missing <- 0.20
powers <- c(1:10, seq(12, 20, by = 2))
fallback_power <- 6

min_module_size <- 15
merge_cut_height <- 0.15
network_type <- "unsigned"
tom_type <- "unsigned"
module_preservation_permutations <- 200
module_preservation_seed <- 20260722
rerun_module_preservation <- FALSE

dir_wgcna <- fs::path(dir_temp_data, "WGCNA")
dir_manuscript <- fs::path(dir_project, "4_manuscripts")
fs::dir_create(dir_wgcna)
fs::dir_create(dir_manuscript)
```

Parameter choices:

- `max_protein_missing = 0.20`: removes proteins with more than 20% missing values before imputation, so modules are not driven by sparse measurements.
- `powers`: candidate soft-thresholding powers used by `pickSoftThreshold()` to choose an approximate scale-free topology.
- `fallback_power = 6`: used only if `pickSoftThreshold()` does not return a power; 6 is a moderate default for WGCNA networks.
- `min_module_size = 15`: allows smaller protein modules, which is reasonable for proteomics where the measured feature set is much smaller than transcriptomics.
- `merge_cut_height = 0.15`: merges modules with highly similar eigengenes, corresponding to eigengene correlation above about 0.85.
- `network_type = "unsigned"` and `tom_type = "unsigned"`: groups proteins using absolute correlation, so strongly positive and strongly negative connections can contribute to the same module.
- `reassignThreshold = 0`: keeps the initial dynamic tree cut assignments stable rather than reassigning borderline proteins.
- `pamRespectsDendro = FALSE`: allows PAM refinement after the tree cut, a common WGCNA setting for blockwise modules.
- `module_preservation_permutations = 200`: uses permutation testing to evaluate whether SOLAR module co-expression structure is preserved in CHS and BHRC. Increase this value for final publication-grade inference if runtime is acceptable.
- `rerun_module_preservation = FALSE`: loads the saved SOLAR preservation RDS when available, avoiding a repeat of the permutation step. Set to `TRUE` to recompute preservation results.

# Data Preparation

```{r proteomics-data}
proteomics_l <- readRDS(fs::path(
  dir_data,
  "ShARP_BHRC_Shared",
  "ShARP_BHRC_Data_Shared",
  "OLINK_proteomics_bridged_BHRC_SOLAR_CHS.rds"
))

overlap_proteins <- proteomics_l[cohort_names] %>%
  purrr::map(~ setdiff(names(.x), proteomics_metadata_cols)) %>%
  purrr::reduce(intersect)

proteomics_l <- proteomics_l[c(cohort_names, "protein_annotation")]

proteomics_l[cohort_names] <- proteomics_l[cohort_names] %>%
  purrr::map(
    ~ .x %>%
      dplyr::select(
        tidyselect::any_of(proteomics_metadata_cols),
        tidyselect::all_of(overlap_proteins)
      )
  )

proteomics_l$protein_annotation <- proteomics_l$protein_annotation[cohort_names] %>%
  purrr::map(~ .x %>% dplyr::filter(meta_id %in% overlap_proteins))

make_cohort_expr <- function(df, cohort_name) {
  df %>%
    dplyr::select(LABID, tidyselect::all_of(overlap_proteins)) %>%
    dplyr::mutate(
      cohort = cohort_name,
      sample_id = paste(cohort_name, LABID, sep = "_")
    ) %>%
    dplyr::select(sample_id, cohort, tidyselect::all_of(overlap_proteins))
}

SOLAR_expr_raw <- make_cohort_expr(proteomics_l$SOLAR, "SOLAR")
CHS_expr_raw <- make_cohort_expr(proteomics_l$CHS, "CHS")
BHRC_expr_raw <- make_cohort_expr(proteomics_l$BHRC, "BHRC")

pooled_expr_raw <- dplyr::bind_rows(
  SOLAR_expr_raw,
  CHS_expr_raw,
  BHRC_expr_raw
)

cat("Overlapping proteins across SOLAR, CHS, and BHRC:", length(overlap_proteins), "\n")
print(table(pooled_expr_raw$cohort))
```

# WGCNA Helper Functions

```{r wgcna-functions}
WGCNA::allowWGCNAThreads()
options(stringsAsFactors = FALSE)

cor <- function(
  x,
  y = NULL,
  use = "pairwise.complete.obs",
  method = "pearson",
  weights.x = NULL,
  weights.y = NULL,
  cosine = FALSE,
  ...
) {
  stats::cor(x, y, use = use, method = method)
}

prepare_wgcna_expr <- function(df,
                               protein_names,
                               sample_cols = c("sample_id", "cohort"),
                               max_missing = 0.20) {
  sample_metadata <- df %>%
    dplyr::select(tidyselect::all_of(sample_cols))

  expr0 <- df %>%
    dplyr::select(tidyselect::all_of(protein_names)) %>%
    dplyr::mutate(dplyr::across(tidyselect::everything(), as.numeric)) %>%
    as.data.frame()

  keep_samples <- !apply(is.na(expr0), 1, all)
  expr0 <- expr0[keep_samples, , drop = FALSE]
  sample_metadata <- sample_metadata[keep_samples, , drop = FALSE]

  protein_missing <- colMeans(is.na(expr0))
  keep_missing <- names(protein_missing)[protein_missing <= max_missing]
  expr1 <- expr0[, keep_missing, drop = FALSE]

  protein_medians <- vapply(expr1, median, numeric(1), na.rm = TRUE)
  expr1 <- as.data.frame(
    Map(
      function(x, med) {
        x[is.na(x)] <- med
        x
      },
      expr1,
      protein_medians
    )
  )

  protein_var <- vapply(expr1, var, numeric(1), na.rm = TRUE)
  keep_var <- names(protein_var)[protein_var > 0]
  expr2 <- expr1[, keep_var, drop = FALSE]

  protein_center <- vapply(expr2, mean, numeric(1), na.rm = TRUE)
  protein_scale <- vapply(expr2, stats::sd, numeric(1), na.rm = TRUE)

  datExpr <- sweep(expr2, 2, protein_center, "-")
  datExpr <- sweep(datExpr, 2, protein_scale, "/")
  datExpr <- as.data.frame(datExpr)
  rownames(datExpr) <- make.unique(sample_metadata$sample_id)

  gsg <- WGCNA::goodSamplesGenes(datExpr, verbose = 3)

  if (!gsg$allOK) {
    good_proteins <- colnames(datExpr)[gsg$goodGenes]

    datExpr <- datExpr[gsg$goodSamples, gsg$goodGenes, drop = FALSE]
    sample_metadata <- sample_metadata[gsg$goodSamples, , drop = FALSE]
    protein_center <- protein_center[good_proteins]
    protein_scale <- protein_scale[good_proteins]
    protein_medians <- protein_medians[good_proteins]
  }

  list(
    datExpr = datExpr,
    sample_metadata = sample_metadata,
    protein_center = protein_center[colnames(datExpr)],
    protein_scale = protein_scale[colnames(datExpr)],
    protein_medians = protein_medians[colnames(datExpr)]
  )
}

prepare_reference_scaled_expr <- function(target_df,
                                          reference_prep,
                                          protein_names,
                                          sample_cols = c("sample_id", "cohort")) {
  target_meta <- target_df %>%
    dplyr::select(tidyselect::all_of(sample_cols))

  target_proteins <- protein_names %>%
    intersect(names(target_df)) %>%
    intersect(names(reference_prep$protein_center)) %>%
    intersect(names(reference_prep$protein_scale)) %>%
    intersect(names(reference_prep$protein_medians))

  if (length(target_proteins) == 0) {
    stop("No target proteins are available for SOLAR-reference scaling.")
  }

  target_expr <- target_df %>%
    dplyr::select(tidyselect::all_of(target_proteins)) %>%
    dplyr::mutate(dplyr::across(tidyselect::everything(), as.numeric)) %>%
    as.data.frame()

  for (protein in target_proteins) {
    missing_idx <- is.na(target_expr[[protein]])
    if (any(missing_idx)) {
      target_expr[[protein]][missing_idx] <- reference_prep$protein_medians[[protein]]
    }
  }

  target_expr <- sweep(
    target_expr,
    2,
    reference_prep$protein_center[target_proteins],
    "-"
  )
  target_expr <- sweep(
    target_expr,
    2,
    reference_prep$protein_scale[target_proteins],
    "/"
  )
  target_expr <- as.data.frame(target_expr)
  rownames(target_expr) <- make.unique(target_meta$sample_id)

  list(
    datExpr = target_expr,
    sample_metadata = target_meta
  )
}

estimate_reference_module_loadings <- function(reference_datExpr,
                                               module_colors,
                                               reference_MEs) {
  modules <- setdiff(sort(unique(module_colors)), "grey")

  purrr::map_dfr(
    modules,
    function(module_name) {
      module_proteins <- names(module_colors)[module_colors == module_name]
      module_expr <- as.matrix(reference_datExpr[, module_proteins, drop = FALSE])

      module_pca <- stats::prcomp(
        module_expr,
        center = FALSE,
        scale. = FALSE
      )

      loading <- module_pca$rotation[, 1]
      score <- as.numeric(module_expr %*% loading)
      eigengene_name <- paste0("ME", module_name)

      if (eigengene_name %in% names(reference_MEs)) {
        sign_cor <- suppressWarnings(stats::cor(score, reference_MEs[[eigengene_name]], use = "pairwise.complete.obs"))
        if (!is.na(sign_cor) && sign_cor < 0) {
          loading <- -loading
          score <- -score
        }
      }

      tibble::tibble(
        module = module_name,
        eigengene = eigengene_name,
        protein = names(loading),
        loading = as.numeric(loading),
        solar_score_cor = if (eigengene_name %in% names(reference_MEs)) {
          suppressWarnings(stats::cor(score, reference_MEs[[eigengene_name]], use = "pairwise.complete.obs"))
        } else {
          NA_real_
        }
      )
    }
  )
}

project_module_eigengenes <- function(target_df,
                                      reference_wgcna,
                                      reference_prep,
                                      cohort_name,
                                      protein_names) {
  module_df <- reference_wgcna$module_df %>%
    dplyr::filter(module != "grey")

  loading_df <- reference_wgcna$module_loadings

  if (is.null(loading_df)) {
    stop("The reference WGCNA object does not contain fixed SOLAR module loadings.")
  }

  projection_proteins <- loading_df$protein %>%
    intersect(protein_names) %>%
    intersect(names(target_df))

  if (length(projection_proteins) == 0) {
    stop("No non-grey SOLAR module proteins are available for projection.")
  }

  target_prep <- prepare_reference_scaled_expr(
    target_df = target_df,
    reference_prep = reference_prep,
    protein_names = projection_proteins
  )
  target_expr <- target_prep$datExpr
  target_meta <- target_prep$sample_metadata

  projected_ME_list <- loading_df %>%
    dplyr::filter(protein %in% projection_proteins) %>%
    split(.$eigengene) %>%
    purrr::map(
      function(module_loadings) {
        proteins <- module_loadings$protein
        loadings <- module_loadings$loading
        names(loadings) <- proteins

        as.numeric(as.matrix(target_expr[, proteins, drop = FALSE]) %*% loadings[proteins])
      }
    )

  projected_MEs <- as.data.frame(projected_ME_list, check.names = FALSE)

  reference_me_names <- setdiff(names(reference_wgcna$MEs), "MEgrey")
  projected_MEs <- projected_MEs[, intersect(reference_me_names, names(projected_MEs)), drop = FALSE]

  dplyr::bind_cols(
    target_meta,
    tibble::as_tibble(projected_MEs)
  ) %>%
    dplyr::mutate(
      projected_from = "SOLAR",
      projection_method = "fixed_SOLAR_PC1_loadings"
    )
}
```

# Overall WGCNA Across Cohorts

## WGCNA model

```{r pooled-wgcna}
pooled_prep <- prepare_wgcna_expr(
  df = pooled_expr_raw,
  protein_names = overlap_proteins,
  sample_cols = c("sample_id", "cohort"),
  max_missing = max_protein_missing
)

cat("Pooled WGCNA matrix dimensions:", dim(pooled_prep$datExpr), "\n")

pooled_sft <- WGCNA::pickSoftThreshold(
  pooled_prep$datExpr,
  powerVector = powers,
  networkType = network_type,
  verbose = 5
)

pooled_soft_power <- pooled_sft$powerEstimate

if (is.na(pooled_soft_power)) {
  pooled_soft_power <- fallback_power
}

pooled_net <- WGCNA::blockwiseModules(
  pooled_prep$datExpr,
  power = pooled_soft_power,
  networkType = network_type,
  TOMType = tom_type,
  minModuleSize = min_module_size,
  reassignThreshold = 0,
  mergeCutHeight = merge_cut_height,
  numericLabels = TRUE,
  pamRespectsDendro = FALSE,
  verbose = 3
)

pooled_module_colors <- WGCNA::labels2colors(pooled_net$colors)
names(pooled_module_colors) <- colnames(pooled_prep$datExpr)
pooled_MEs0 <- WGCNA::moduleEigengenes(
  pooled_prep$datExpr,
  colors = pooled_module_colors
)$eigengenes
pooled_MEs0 <- WGCNA::orderMEs(pooled_MEs0)

pooled_module_df <- tibble::tibble(
  protein = colnames(pooled_prep$datExpr),
  module = pooled_module_colors
)

pooled_wgcna <- list(
  analysis = "pooled",
  datExpr = pooled_prep$datExpr,
  soft_power = pooled_soft_power,
  soft_threshold_fit = tibble::as_tibble(pooled_sft$fitIndices),
  net = pooled_net,
  module_colors = pooled_module_colors,
  MEs = pooled_MEs0,
  module_df = pooled_module_df
)

pooled_MEs <- dplyr::bind_cols(
  pooled_prep$sample_metadata,
  tibble::as_tibble(pooled_wgcna$MEs)
)

table(pooled_wgcna$module_colors)
```

## Module overview

```{r pooled-module-overview}
pooled_module_overview <- pooled_wgcna$module_df %>%
  dplyr::count(module, name = "n_proteins") %>%
  dplyr::arrange(dplyr::desc(n_proteins))

pooled_module_overview %>%
  dplyr::summarise(
    n_modules = sum(module != "grey"),
    n_unassigned_proteins = sum(n_proteins[module == "grey"]),
    n_module_proteins = sum(n_proteins[module != "grey"])
  )

pooled_module_overview %>%
  kbl(
    caption = "Number of proteins assigned to each pooled WGCNA module."
  ) %>%
  kable_paper("hover", full_width = T)
```

# SOLAR Reference WGCNA

```{r solar-reference-wgcna}
SOLAR_prep <- prepare_wgcna_expr(
  df = SOLAR_expr_raw,
  protein_names = overlap_proteins,
  sample_cols = c("sample_id", "cohort"),
  max_missing = max_protein_missing
)

cat("SOLAR reference WGCNA matrix dimensions:", dim(SOLAR_prep$datExpr), "\n")

SOLAR_sft <- WGCNA::pickSoftThreshold(
  SOLAR_prep$datExpr,
  powerVector = powers,
  networkType = network_type,
  verbose = 5
)

SOLAR_soft_power <- SOLAR_sft$powerEstimate

if (is.na(SOLAR_soft_power)) {
  SOLAR_soft_power <- fallback_power
}

SOLAR_net <- WGCNA::blockwiseModules(
  SOLAR_prep$datExpr,
  power = SOLAR_soft_power,
  networkType = network_type,
  TOMType = tom_type,
  minModuleSize = min_module_size,
  reassignThreshold = 0,
  mergeCutHeight = merge_cut_height,
  numericLabels = TRUE,
  pamRespectsDendro = FALSE,
  verbose = 3
)

SOLAR_module_colors <- WGCNA::labels2colors(SOLAR_net$colors)
names(SOLAR_module_colors) <- colnames(SOLAR_prep$datExpr)
SOLAR_MEs0 <- WGCNA::moduleEigengenes(
  SOLAR_prep$datExpr,
  colors = SOLAR_module_colors
)$eigengenes
SOLAR_MEs0 <- WGCNA::orderMEs(SOLAR_MEs0)

SOLAR_module_df <- tibble::tibble(
  protein = colnames(SOLAR_prep$datExpr),
  module = SOLAR_module_colors
)

SOLAR_module_loadings <- estimate_reference_module_loadings(
  reference_datExpr = SOLAR_prep$datExpr,
  module_colors = SOLAR_module_colors,
  reference_MEs = SOLAR_MEs0
)

SOLAR_wgcna <- list(
  analysis = "SOLAR_reference",
  datExpr = SOLAR_prep$datExpr,
  soft_power = SOLAR_soft_power,
  soft_threshold_fit = tibble::as_tibble(SOLAR_sft$fitIndices),
  net = SOLAR_net,
  module_colors = SOLAR_module_colors,
  MEs = SOLAR_MEs0,
  module_df = SOLAR_module_df,
  module_loadings = SOLAR_module_loadings
)

SOLAR_MEs <- dplyr::bind_cols(
  SOLAR_prep$sample_metadata,
  tibble::as_tibble(SOLAR_wgcna$MEs)
)

table(SOLAR_wgcna$module_colors)
```

## Module overview

```{r solar-module-overview}
SOLAR_module_overview <- SOLAR_wgcna$module_df %>%
  dplyr::count(module, name = "n_proteins") %>%
  dplyr::arrange(dplyr::desc(n_proteins))

SOLAR_module_overview %>%
  dplyr::summarise(
    n_modules = sum(module != "grey"),
    n_unassigned_proteins = sum(n_proteins[module == "grey"]),
    n_module_proteins = sum(n_proteins[module != "grey"])
  )

SOLAR_module_overview %>%
  kbl(
    caption = "Number of proteins assigned to each SOLAR reference WGCNA module."
  ) %>%
  kable_paper("hover")
```

# SOLAR Module Preservation in CHS and BHRC

Module preservation tests whether the SOLAR-derived module co-expression structure is retained in CHS and BHRC.
Unlike eigengene projection, this step does not create participant-level scores.
Instead, it evaluates whether proteins assigned to each SOLAR module remain similarly connected in the target cohorts.
The main statistics are `Zsummary.pres` and `medianRank.pres`: larger `Zsummary.pres` and smaller `medianRank.pres` indicate stronger preservation.

```{r solar-module-preservation}
preservation_proteins <- colnames(SOLAR_prep$datExpr) %>%
  intersect(names(CHS_expr_raw)) %>%
  intersect(names(BHRC_expr_raw))

CHS_preservation_prep <- prepare_reference_scaled_expr(
  target_df = CHS_expr_raw,
  reference_prep = SOLAR_prep,
  protein_names = preservation_proteins
)

BHRC_preservation_prep <- prepare_reference_scaled_expr(
  target_df = BHRC_expr_raw,
  reference_prep = SOLAR_prep,
  protein_names = preservation_proteins
)

preservation_modules <- SOLAR_wgcna$module_colors[preservation_proteins]

extract_preservation_table <- function(module_preservation, test_cohort) {
  ref_name <- "ref.SOLAR"
  test_name <- paste0("inColumnsAlsoPresentIn.", test_cohort)

  z_tbl <- module_preservation$preservation$Z[[ref_name]][[test_name]] %>%
    as.data.frame() %>%
    tibble::rownames_to_column("module") %>%
    tibble::as_tibble()

  observed_tbl <- module_preservation$preservation$observed[[ref_name]][[test_name]] %>%
    as.data.frame() %>%
    tibble::rownames_to_column("module") %>%
    tibble::as_tibble()

  z_cols <- intersect(
    c("module", "moduleSize", "Zsummary.pres", "Zdensity.pres", "Zconnectivity.pres"),
    names(z_tbl)
  )
  observed_cols <- intersect(
    c("module", "medianRank.pres"),
    names(observed_tbl)
  )

  z_tbl %>%
    dplyr::select(tidyselect::all_of(z_cols)) %>%
    dplyr::left_join(
      observed_tbl %>%
        dplyr::select(tidyselect::all_of(observed_cols)),
      by = "module"
    ) %>%
    dplyr::mutate(
      cohort = test_cohort,
      preservation_interpretation = dplyr::case_when(
        !is.na(Zsummary.pres) & Zsummary.pres > 10 ~ "Strong preservation",
        !is.na(Zsummary.pres) & Zsummary.pres >= 2 ~ "Moderate/weak preservation",
        !is.na(Zsummary.pres) ~ "No evidence of preservation",
        TRUE ~ NA_character_
      )
    ) %>%
    dplyr::select(cohort, tidyselect::everything()) %>%
    dplyr::arrange(module)
}

module_preservation_rds <- fs::path(dir_wgcna, "SOLAR_module_preservation.rds")

if (fs::file_exists(module_preservation_rds) && !rerun_module_preservation) {
  SOLAR_module_preservation_saved <- readRDS(module_preservation_rds)
  SOLAR_module_preservation <- SOLAR_module_preservation_saved$module_preservation

  if ("module_preservation_summary" %in% names(SOLAR_module_preservation_saved)) {
    SOLAR_module_preservation_summary <- SOLAR_module_preservation_saved$module_preservation_summary
  } else {
    SOLAR_module_preservation_summary <- dplyr::bind_rows(
      extract_preservation_table(SOLAR_module_preservation, "CHS"),
      extract_preservation_table(SOLAR_module_preservation, "BHRC")
    )
  }
} else {
  multi_expr <- list(
    SOLAR = list(data = SOLAR_prep$datExpr[, preservation_proteins, drop = FALSE]),
    CHS = list(data = CHS_preservation_prep$datExpr[, preservation_proteins, drop = FALSE]),
    BHRC = list(data = BHRC_preservation_prep$datExpr[, preservation_proteins, drop = FALSE])
  )

  multi_color <- list(
    SOLAR = preservation_modules
  )

  set.seed(module_preservation_seed)
  SOLAR_module_preservation <- WGCNA::modulePreservation(
    multiData = multi_expr,
    multiColor = multi_color,
    referenceNetworks = 1,
    nPermutations = module_preservation_permutations,
    networkType = network_type,
    verbose = 3,
    randomSeed = module_preservation_seed
  )

  SOLAR_module_preservation_summary <- dplyr::bind_rows(
    extract_preservation_table(SOLAR_module_preservation, "CHS"),
    extract_preservation_table(SOLAR_module_preservation, "BHRC")
  )
}

SOLAR_module_preservation_summary <- SOLAR_module_preservation_summary %>%
  dplyr::filter(module %in% setdiff(unique(SOLAR_wgcna$module_df$module), "grey"))

SOLAR_module_preservation_summary %>%
  dplyr::arrange(desc(cohort), desc(moduleSize)) |>
  kbl(
    caption = "Preservation of SOLAR reference WGCNA modules in CHS and BHRC."
  ) %>%
  kable_paper("hover")
```

## Supplemental Table: Module Preservation

```{r supplemental-module-preservation-table}
module_preservation_supplemental_table <- SOLAR_module_preservation_summary %>%
  dplyr::transmute(
    `Validation cohort` = cohort,
    `SOLAR module` = module,
    `Module size` = moduleSize,
    `Zsummary preservation` = round(Zsummary.pres, 2),
    `Zdensity preservation` = round(Zdensity.pres, 2),
    `Zconnectivity preservation` = round(Zconnectivity.pres, 2),
    `Median rank preservation` = round(medianRank.pres, 2),
    `Interpretation` = preservation_interpretation
  ) %>%
  dplyr::arrange(`Validation cohort`, dplyr::desc(`Module size`), `SOLAR module`)

# readr::write_csv(
#   module_preservation_supplemental_table,
#   fs::path(dir_manuscript, "Supplemental_Table_Module_Preservation.csv")
# )

module_preservation_supplemental_table %>%
  kbl(
    caption = "Supplemental Table. Preservation of SOLAR reference WGCNA protein modules in CHS and BHRC."
  ) %>%
  kable_paper("hover", full_width = FALSE)
```

Columns in `SOLAR_module_preservation_summary`:

- `Zsummary.pres`: composite preservation Z statistic. Values greater than 10 indicate strong preservation, values from 2 to 10 indicate moderate or weak evidence of preservation, and values below 2 indicate no clear evidence of preservation.
- `Zdensity.pres`: preservation statistic for whether proteins in the module remain densely connected in the target cohort.
- `Zconnectivity.pres`: preservation statistic for whether the within-module connectivity pattern is retained in the target cohort.
- `medianRank.pres`: relative preservation rank across modules. Smaller values indicate stronger preservation. This statistic is useful because it is less dependent on module size than `Zsummary.pres`.
- `preservation_interpretation`: text interpretation derived from `Zsummary.pres`.

```{r save-solar-module-preservation}
saveRDS(
  list(
    module_preservation = SOLAR_module_preservation,
    module_preservation_summary = SOLAR_module_preservation_summary,
    module_preservation_supplemental_table = module_preservation_supplemental_table
  ),
  fs::path(dir_wgcna, "SOLAR_module_preservation.rds")
)
```

# SOLAR Module Projection to CHS and BHRC

The SOLAR reference modules are projected into CHS and BHRC with fixed SOLAR eigengene loadings.
For each non-grey SOLAR module, the first principal component loading vector is estimated once in the standardized SOLAR protein matrix and aligned to the direction of the original SOLAR WGCNA eigengene.
CHS and BHRC protein abundances are then imputed with SOLAR medians, standardized using SOLAR means and standard deviations, and multiplied by the fixed SOLAR loading vectors.
Thus, target-cohort module scores use SOLAR module membership and SOLAR eigengene weights rather than recomputing module eigengenes separately inside each target cohort.

```{r solar-module-projection}
CHS_projected_MEs <- project_module_eigengenes(
  target_df = CHS_expr_raw,
  reference_wgcna = SOLAR_wgcna,
  reference_prep = SOLAR_prep,
  cohort_name = "CHS",
  protein_names = overlap_proteins
)

BHRC_projected_MEs <- project_module_eigengenes(
  target_df = BHRC_expr_raw,
  reference_wgcna = SOLAR_wgcna,
  reference_prep = SOLAR_prep,
  cohort_name = "BHRC",
  protein_names = overlap_proteins
)

projected_MEs <- dplyr::bind_rows(
  CHS_projected_MEs,
  BHRC_projected_MEs
)

print(table(projected_MEs$cohort))
```

# Saved WGCNA Models

```{r save-wgcna-model}
WGCNA_parameters <- list(
  cohort_names = cohort_names,
  max_protein_missing = max_protein_missing,
  powers = powers,
  fallback_power = fallback_power,
  min_module_size = min_module_size,
  merge_cut_height = merge_cut_height,
  network_type = network_type,
  tom_type = tom_type,
  module_preservation_permutations = module_preservation_permutations,
  module_preservation_seed = module_preservation_seed
)

pooled_WGCNA_model <- list(
  parameters = WGCNA_parameters,
  prep = pooled_prep,
  wgcna = pooled_wgcna,
  module_overview = pooled_module_overview,
  module_eigengenes = pooled_MEs
)

SOLAR_WGCNA_model <- list(
  parameters = WGCNA_parameters,
  reference = list(
    prep = SOLAR_prep,
    wgcna = SOLAR_wgcna,
    module_overview = SOLAR_module_overview,
    module_eigengenes = SOLAR_MEs
  ),
  preservation = list(
    module_preservation = SOLAR_module_preservation,
    module_preservation_summary = SOLAR_module_preservation_summary,
    module_preservation_supplemental_table = module_preservation_supplemental_table,
    CHS_preservation_prep = CHS_preservation_prep,
    BHRC_preservation_prep = BHRC_preservation_prep
  ),
  projection = list(
    CHS_projected_MEs = CHS_projected_MEs,
    BHRC_projected_MEs = BHRC_projected_MEs,
    projected_MEs = projected_MEs
  )
)

saveRDS(pooled_WGCNA_model, fs::path(dir_wgcna, "pooled_WGCNA_model.rds"))
saveRDS(SOLAR_WGCNA_model, fs::path(dir_wgcna, "SOLAR_WGCNA_model.rds"))
```
