---
title: "SOL, CHS, and BHRC PFAS PWAS"
author: "Zhenjiang Li"
date: today
toc: true
toc_expand: 2
format:
  html:
    code-fold: true
    html-math-method: katex
    fontsize: 16pt
    grid:
      sidebar-width: 0px
      body-width: 1500px
      margin-width: 500px
      gutter-width: 1.5rem
embed-resources: true
editor: visual
editor_options:
  chunk_output_type: console
---

## Environment Setup {.hidden}

```{=html}
<style type="text/css">
caption, .table-caption {
  text-align: left;
}
</style>
```

```{r setup}
#| echo: false

options(scipen=999)

library(fs)
library(here)
i_am("1_scripts/!analytic_workflow_pfas_PWAS.qmd")

source(here::here("1_scripts", "!libraries.R"))
source(here::here("1_scripts", "!directories.R"))
source(here::here("1_scripts", "!functions.R"))

options(knitr.table.format = "html")
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

```{r}
#| label: shorten-long-path
#| include: false

create_shortcut <- function(path, short_cut, drive)
{
  return(
    sub(gsub("\\)", "\\\\)", gsub("\\(", "\\\\(", short_cut)), paste0(drive, ":"), path))
}

if(nchar(dir_data) > 80)
{
  system("subst x: /D")
  short_cut <- dir_home |> dirname()
  system(paste0("subst x: \"", short_cut, "\""))

  dir_objects_names <- ls(pattern = "^dir_")

  dir_list <- mget(dir_objects_names)

  dir_list_shortened <- lapply(
    dir_list,
    function(y) create_shortcut(y, short_cut, "x")
  )
  list2env(dir_list_shortened, envir = .GlobalEnv)
}

rm(list = intersect(
  c("dir_objects_names", "short_cut", "dir_list", "dir_list_shortened"),
  ls()
))
```

# Parameters

```{r}
SOLAR_pfas_names <- c("pfos_log2", "pfoa_log2", "pfna_log2", "pfda_log2", "pfhxs_log2", "pfhps_log2")
CHS_pfas_names <- c("ma_pfos_log2", "ma_pfoa_log2", "ma_pfna_log2", "ma_pfda_log2", "ma_pfhxs_log2", "ma_pfhps_log2")
BHRC_pfas_names <- c("PFOS_det60_imputed_log2", "PFOA_det60_imputed_log2", "PFNA_det60_imputed_log2", "PFDA_det60_imputed_log2", "PFHxS_det60_imputed_log2", "PFHpS_det60_imputed_log2")
PFAS_abbrev_names <- c("PFOS", "PFOA", "PFNA", "PFDA", "PFHxS", "PFHpS")

SOLAR_covars <- c("age", "sex", "bmiz", "tanner", "wave", "edu_house2")
CHS_covars <- c("ma_age", "ma_sex", "ma_hisp", "ma_parental_edu", "ma_bmi")
BHRC_raw_covars <- c("AGE_AT_VISIT", "BMI1", "GENDER", "YEARS_EDU", "ADA2010_DM")
BHRC_covars <- c("AGE_AT_VISIT", "BMI1", "GENDER", "edu_3level", "ADA2010_DM")

qgcomp_q <- 4
dir_pwas <- fs::path(dir_temp_data, "PFAS_PWAS")
fs::dir_create(dir_pwas)
```

# Data Preparation

## Bridged proteomics

```{r}
proteomics_l <- readRDS(fs::path(
  dir_data,
  "ShARP_BHRC_Shared",
  "ShARP_BHRC_Data_Shared",
  "OLINK_proteomics_bridged_BHRC_SOLAR_CHS.rds"
  ))

cohort_names <- c("SOLAR", "CHS", "BHRC")
proteomics_metadata_cols <- c("LABID", "study", "Batch")

overlap_proteins <- proteomics_l[cohort_names] %>%
  purrr::map(~ setdiff(names(.x), proteomics_metadata_cols)) %>%
  purrr::reduce(intersect)

proteomics_l <- proteomics_l[c(cohort_names, "protein_annotation")]

proteomics_l[cohort_names] <- proteomics_l[cohort_names] %>%
  purrr::map(~ .x %>% dplyr::select(tidyselect::any_of(proteomics_metadata_cols), tidyselect::all_of(overlap_proteins)))

proteomics_l$protein_annotation <- proteomics_l$protein_annotation[cohort_names] %>%
  purrr::map(~ .x %>% dplyr::filter(meta_id %in% overlap_proteins))

SOL_prote <- proteomics_l$SOLAR %>%
  dplyr::mutate(id = gsub("-", "_", tolower(LABID))) %>%
  dplyr::select(id, tidyselect::all_of(overlap_proteins))

CHS_prote <- proteomics_l$CHS %>%
  dplyr::rename(SampleID = LABID) %>%
  dplyr::select(SampleID, tidyselect::all_of(overlap_proteins))

BHRC_prote <- proteomics_l$BHRC %>%
  dplyr::select(LABID, tidyselect::all_of(overlap_proteins))

proteomics_property <- proteomics_l$protein_annotation$SOLAR %>%
  dplyr::transmute(
    OlinkID = meta_id,
    UniProt = meta_Uniprot,
    Assay = meta_id,
    Protein = meta_Assay,
    Panel = NA_character_
  )

cat("Overlapping proteins across SOLAR, CHS, and BHRC:", length(overlap_proteins), "\n")
names(proteomics_l)
```

## SOLAR

```{r solar-data-prep}
load(fs::path(
  dir_data,
  "ShARP_SOLAR_Shared",
  "ShARP_SOLAR_Data_Shared",
  "Solar_Analysis_20251003.Rda"
))
SOLAR <- analysis_df
rm(analysis_df)

SOLAR_visit1 <- SOLAR %>%
  dplyr::group_by(id) %>%
  dplyr::filter(visit == min(visit)) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    age = as.numeric(age),
    bmiz = as.numeric(bmiz),
    edu_house2 = dplyr::recode(
      edu_house,
      "Missing" = "Missing",
      "Did not graduate high school" = "< High School",
      "High school graduate" = "High School Grad",
      "Partial college (at least one year) or specialized training" = "> High School",
      "Completed college/university" = "> High School",
      "Graduate professional training (graduate degree)" = "> High School"
    ),
    dplyr::across(tidyselect::any_of(c("sex", "tanner", "wave", "edu_house2")), as.factor)
  ) %>%
  dplyr::select(
    id,
    tidyselect::all_of(SOLAR_pfas_names),
    tidyselect::all_of(SOLAR_covars),
    bmi_og
  )

colSums(is.na(SOLAR_visit1))

SOLAR_working_df <- SOL_prote %>%
  dplyr::left_join(SOLAR_visit1, by = "id")

colSums(is.na(SOLAR_working_df))
dim(SOLAR_working_df)
```

## CHS

```{r chs-data-prep}
CHS <- readRDS(fs::path(
  dir_data,
  "ShARP_CHS_Shared",
  "ShARP_CHS_Data_Shared",
  "CHS MetaAir MetaChem Cleaned Redcap and Exposures Outcome Data long V5.rds"))

id_link <- readxl::read_xlsx(
  fs::path(
  here::here() |> dirname() |> dirname() |> dirname() |> dirname() |> dirname(),
  "Chatzi Projects and Analyses", "Env Chem SOL-CHS",
  "Lab Analysis and sample shipments", "Metachem_sample_inventory_master_040522.xlsx"),
  sheet = "Plasma"
)

id_link_simp <- id_link %>%
  dplyr::mutate(
    SampleID = `Plasma Aliquot ID`,
    id = `Sample Name`
  ) %>%
  dplyr::mutate(
    id = as.character(id)
  ) %>%
  dplyr::select(SampleID, id)

CHS_prote <- CHS_prote |>
  dplyr::left_join(
    id_link_simp,
    by = c("SampleID")
  ) %>%
  dplyr::select(id, tidyselect::everything(), -SampleID)
duplicated(CHS_prote$id) |> sum() # all unique IDs

CHS_baseline <- CHS %>%
  dplyr::filter(study == "MetaAir") %>%
  dplyr::select(
    id,
    tidyselect::all_of(CHS_pfas_names),
    tidyselect::all_of(CHS_covars)
  ) %>%
  dplyr::mutate(
    ma_age = as.numeric(ma_age),
    ma_bmi = as.numeric(ma_bmi),
    ma_hisp = as.factor(ma_hisp),
    ma_parental_edu = dplyr::na_if(ma_parental_edu, -1),
    ma_parental_edu = as.factor(ma_parental_edu),
    ma_sex = as.factor(ma_sex)
  )

colSums(is.na(CHS_baseline))

CHS_analysis_df <- CHS_baseline %>%
  dplyr::filter(complete.cases(.))

CHS_working_df <- dplyr::left_join(
  CHS_prote,
  CHS_analysis_df,
  by = "id"
)

colSums(is.na(CHS_working_df))
dim(CHS_working_df)
```

## BHRC

```{r}
BHRC <- readRDS(
  fs::path(
    dir_data,
    "ShARP_BHRC_Shared",
    "ShARP_BHRC_Data_Shared",
    "CCHC_usc250_long_12172025.rds"
  )
) |>
  dplyr::filter(visit %in% c(1,2))
  # dplyr::filter(visit %in% c(1))

BHRC_pfas <- readRDS(
  fs::path(
    dir_data,
    "ShARP_BHRC_Shared",
    "ShARP_BHRC_Data_Shared",
    "PFAS_CCHC_three_sets.rds"
  )
)

BHRC_analysis_df <- BHRC %>%
  dplyr::filter(visit == 1) %>%
  dplyr::select(
    RRID,
    LABID,
    visit,
    tidyselect::all_of(BHRC_raw_covars)
  ) %>%
  dplyr::left_join(
    BHRC_pfas %>%
      dplyr::select(RRID, tidyselect::all_of(BHRC_pfas_names)),
    by = "RRID"
  ) %>%
  dplyr::mutate(
    AGE_AT_VISIT = as.numeric(AGE_AT_VISIT),
    BMI1 = as.numeric(BMI1),
    GENDER = as.factor(GENDER),
    YEARS_EDU_code = code_from_years_edu(YEARS_EDU),
    edu_3level = bhrc_edu_3level(YEARS_EDU_code),
    ADA2010_DM = as.factor(ADA2010_DM)
  ) %>%
  dplyr::filter(complete.cases(dplyr::select(., tidyselect::all_of(c(BHRC_pfas_names, BHRC_covars))))) %>%
  dplyr::arrange(LABID, visit) %>%
  dplyr::distinct(LABID, .keep_all = TRUE)

BHRC_working_df <- BHRC_prote %>%
  dplyr::left_join(BHRC_analysis_df, by = "LABID")

colSums(is.na(BHRC_working_df))
dim(BHRC_working_df)
```

# Table 1

```{r}
lapply(
  list(
    SOLAR_working_df$age,
    CHS_working_df$ma_age,
    BHRC_working_df$AGE_AT_VISIT
  ),
  function(x) range(x, na.rm = T)
)

table1::table1(
  as.formula(paste0("~", paste(SOLAR_covars, collapse = "+"), "+bmi_og")), data = SOLAR_working_df
)
table1::table1(
  as.formula(paste0("~", paste(CHS_covars, collapse = "+"))), data = CHS_working_df
)
table1::table1(
  as.formula(paste0("~", paste(BHRC_covars, collapse = "+"))), data = BHRC_working_df
)

pfas_summary_by_cohort <- function(df, cohort, pfas_names) {
  pfas_name_map <- stats::setNames(pfas_names, PFAS_abbrev_names)

  df %>%
    dplyr::select(tidyselect::all_of(pfas_names)) %>%
    dplyr::rename(tidyselect::all_of(pfas_name_map)) %>%
    tidyr::pivot_longer(
      cols = tidyselect::all_of(PFAS_abbrev_names),
      names_to = "PFAS",
      values_to = "log2_value"
    ) %>%
    dplyr::mutate(
      concentration = 2^log2_value
    ) %>%
    dplyr::group_by(PFAS) %>%
    dplyr::summarise(
      cohort = cohort,
      median = stats::median(concentration, na.rm = TRUE),
      q1 = stats::quantile(concentration, probs = 0.25, na.rm = TRUE, names = FALSE),
      q3 = stats::quantile(concentration, probs = 0.75, na.rm = TRUE, names = FALSE),
      .groups = "drop"
    )
}

pfas_concentration_table1 <- dplyr::bind_rows(
  pfas_summary_by_cohort(SOLAR_working_df, "SOLAR", SOLAR_pfas_names),
  pfas_summary_by_cohort(CHS_working_df, "CHS", CHS_pfas_names),
  pfas_summary_by_cohort(BHRC_working_df, "BHRC", BHRC_pfas_names)
) %>%
  dplyr::mutate(
    PFAS = factor(PFAS, levels = PFAS_abbrev_names),
    median_iqr = sprintf("%.2f (%.2f, %.2f)", median, q1, q3)
  ) %>%
  dplyr::select(PFAS, cohort, median_iqr) %>%
  tidyr::pivot_wider(names_from = cohort, values_from = median_iqr) %>%
  dplyr::arrange(PFAS)

pfas_concentration_table1 %>%
  kableExtra::kbl(
    caption = "Common PFAS concentrations across cohorts, median (Q1, Q3), original concentration scale."
  ) %>%
  kableExtra::kable_styling(full_width = FALSE)

```

# PWAS

This section performs two proteome-wide screening analyses in each cohort.
First, each PFAS compound is tested separately against every protein using `epiomics::owas()` with cohort-specific covariate adjustment.
Second, the six PFAS compounds are tested jointly as a mixture using `epiomics::owas_qgcomp()`.
Only two combined result files are saved because the downstream Praxis-BGM workflow only needs the screening results.

## Analysis Helpers

```{r pwas-helpers}
prepare_pwas_df <- function(df, id_col, pfas_names, covars, protein_names) {
  df %>%
    dplyr::select(
      tidyselect::all_of(id_col),
      tidyselect::all_of(pfas_names),
      tidyselect::all_of(covars),
      tidyselect::all_of(protein_names)
    ) %>%
    dplyr::filter(
      stats::complete.cases(dplyr::select(., tidyselect::all_of(c(pfas_names, covars))))
    )
}

annotate_single_pfas_results <- function(res, cohort, pfas_var, pfas_names) {
  res %>%
    dplyr::mutate(
      cohort = cohort,
      pfas = pfas_var,
      pfas_abbrev = PFAS_abbrev_names[match(pfas_var, pfas_names)],
      adjusted_pval_BH = stats::p.adjust(p_value, method = "BH"),
      adjusted_pval_BY = stats::p.adjust(p_value, method = "BY"),
      estimate_beta = estimate,
      se_beta = se,
      ci_beta_low = estimate_beta - 1.96 * se_beta,
      ci_beta_high = estimate_beta + 1.96 * se_beta
    )
}

standardize_mixture_feature_column <- function(res) {
  if ("feature" %in% names(res) && !"feature_name" %in% names(res)) {
    res <- dplyr::rename(res, feature_name = feature)
  }
  res
}

run_cohort_single_pwas <- function(df, cohort, pfas_names, covars, protein_names) {
  purrr::map_dfr(
    pfas_names,
    function(pfas_var) {
      epiomics::owas(
        df = df,
        var = pfas_var,
        omics = protein_names,
        covars = covars,
        var_exposure_or_outcome = "exposure",
        family = "gaussian"
      ) %>%
        annotate_single_pfas_results(cohort, pfas_var, pfas_names)
    }
  ) %>%
    dplyr::left_join(proteomics_property, by = c("feature_name" = "Assay")) %>%
    dplyr::relocate(
      cohort,
      pfas,
      pfas_abbrev,
      feature_name,
      Protein,
      UniProt,
      OlinkID,
      Panel,
      .before = estimate
    )
}

run_cohort_mixture_pwas <- function(df, cohort, pfas_names, covars, protein_names) {
  epiomics::owas_qgcomp(
    df = df,
    expnms = pfas_names,
    omics = protein_names,
    covars = covars,
    q = qgcomp_q,
    confidence_level = 0.95,
    family = "gaussian"
  ) %>%
    standardize_mixture_feature_column() %>%
    dplyr::mutate(
      cohort = cohort,
      mixture = "PFAS mixture",
      pfas = "PFAS mixture",
      pfas_abbrev = "PFAS mixture",
      adjusted_pval_BH = stats::p.adjust(p_value, method = "BH"),
      adjusted_pval_BY = stats::p.adjust(p_value, method = "BY"),
      estimate_beta = psi,
      ci_beta_low = lcl_psi,
      ci_beta_high = ucl_psi
    ) %>%
    dplyr::left_join(proteomics_property, by = c("feature_name" = "Assay")) %>%
    dplyr::relocate(
      cohort,
      mixture,
      pfas,
      pfas_abbrev,
      feature_name,
      Protein,
      UniProt,
      OlinkID,
      Panel,
      .before = estimate_beta
    )
}
```

## Complete-Case PWAS Inputs

```{r pwas-inputs}
SOLAR_pwas_df <- SOLAR_working_df %>%
  prepare_pwas_df("id", SOLAR_pfas_names, SOLAR_covars, overlap_proteins)

CHS_pwas_df <- CHS_working_df %>%
  prepare_pwas_df("id", CHS_pfas_names, CHS_covars, overlap_proteins)

BHRC_pwas_df <- BHRC_working_df %>%
  prepare_pwas_df("LABID", BHRC_pfas_names, BHRC_covars, overlap_proteins)

tibble::tibble(
  cohort = cohort_names,
  n_participants = c(nrow(SOLAR_pwas_df), nrow(CHS_pwas_df), nrow(BHRC_pwas_df)),
  n_pfas = length(PFAS_abbrev_names),
  n_proteins = length(overlap_proteins)
) %>%
  kableExtra::kbl(caption = "Complete-case PWAS input dimensions.") %>%
  kableExtra::kable_paper("hover", full_width = FALSE)
```

## Single-PFAS PWAS

```{r single-pfas-pwas}
SOLAR_PWAS_results <- run_cohort_single_pwas(
  df = SOLAR_pwas_df,
  cohort = "SOLAR",
  pfas_names = SOLAR_pfas_names,
  covars = SOLAR_covars,
  protein_names = overlap_proteins
)

CHS_PWAS_results <- run_cohort_single_pwas(
  df = CHS_pwas_df,
  cohort = "CHS",
  pfas_names = CHS_pfas_names,
  covars = CHS_covars,
  protein_names = overlap_proteins
)

BHRC_PWAS_results <- run_cohort_single_pwas(
  df = BHRC_pwas_df,
  cohort = "BHRC",
  pfas_names = BHRC_pfas_names,
  covars = BHRC_covars,
  protein_names = overlap_proteins
)

PFAS_single_PWAS_results <- dplyr::bind_rows(
  SOLAR_PWAS_results,
  CHS_PWAS_results,
  BHRC_PWAS_results
)

readr::write_csv(
  PFAS_single_PWAS_results,
  fs::path(dir_pwas, "PFAS_single_PWAS_results.csv")
)

PFAS_single_PWAS_results %>%
  dplyr::group_by(cohort, pfas_abbrev) %>%
  dplyr::summarise(
    n_nominal_p_lt_0.05 = sum(p_value < 0.05, na.rm = TRUE),
    n_BH_p_lt_0.05 = sum(adjusted_pval_BH < 0.05, na.rm = TRUE),
    n_BY_p_lt_0.05 = sum(adjusted_pval_BY < 0.05, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  dplyr::arrange(cohort, pfas_abbrev) %>%
  kableExtra::kbl(caption = "Single-PFAS PWAS significant association counts.") %>%
  kableExtra::kable_paper("hover", full_width = FALSE)
```

## PFAS-Mixture PWAS

```{r mixture-pfas-pwas}
SOLAR_mixture_PWAS_results <- run_cohort_mixture_pwas(
  df = SOLAR_pwas_df,
  cohort = "SOLAR",
  pfas_names = SOLAR_pfas_names,
  covars = SOLAR_covars,
  protein_names = overlap_proteins
)

CHS_mixture_PWAS_results <- run_cohort_mixture_pwas(
  df = CHS_pwas_df,
  cohort = "CHS",
  pfas_names = CHS_pfas_names,
  covars = CHS_covars,
  protein_names = overlap_proteins
)

BHRC_mixture_PWAS_results <- run_cohort_mixture_pwas(
  df = BHRC_pwas_df,
  cohort = "BHRC",
  pfas_names = BHRC_pfas_names,
  covars = BHRC_covars,
  protein_names = overlap_proteins
)

PFAS_mixture_PWAS_results <- dplyr::bind_rows(
  SOLAR_mixture_PWAS_results,
  CHS_mixture_PWAS_results,
  BHRC_mixture_PWAS_results
)

readr::write_csv(
  PFAS_mixture_PWAS_results,
  fs::path(dir_pwas, "PFAS_mixture_qgcomp_PWAS_results.csv")
)

PFAS_mixture_PWAS_results %>%
  dplyr::group_by(cohort, mixture) %>%
  dplyr::summarise(
    n_nominal_p_lt_0.05 = sum(p_value < 0.05, na.rm = TRUE),
    n_BH_p_lt_0.05 = sum(adjusted_pval_BH < 0.05, na.rm = TRUE),
    n_BY_p_lt_0.05 = sum(adjusted_pval_BY < 0.05, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  dplyr::arrange(cohort) %>%
  kableExtra::kbl(caption = "PFAS-mixture qgcomp PWAS significant association counts.") %>%
  kableExtra::kable_paper("hover", full_width = FALSE)
```
