---
title: "PFAS Proteomics Final Validation"
author: "Zhenjiang Li"
date: today
toc: true
toc_expand: 2
format:
  html:
    code-fold: true
    html-math-method: katex
    fontsize: 16pt
    grid:
      sidebar-width: 0px
      body-width: 1500px
      margin-width: 500px
      gutter-width: 1.5rem
embed-resources: true
editor: visual
editor_options:
  chunk_output_type: console
---

## Environment Setup {.hidden}

```{=html}
<style type="text/css">
caption, .table-caption {
  text-align: left;
}
</style>
```

```{r setup}
#| echo: false

options(scipen = 999)

library(fs)
library(here)
i_am("1_scripts/!analytic_workflow_pfas_final_validation.qmd")

source(here::here("1_scripts", "!libraries.R"))
source(here::here("1_scripts", "!directories.R"))
source(here::here("1_scripts", "!functions.R"))

options(knitr.table.format = "html")
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r shorten-long-path}
#| include: false

create_shortcut <- function(path, short_cut, drive) {
  return(
    sub(gsub("\\)", "\\\\)", gsub("\\(", "\\\\(", short_cut)), paste0(drive, ":"), path)
  )
}

if (nchar(dir_data) > 80) {
  system("subst x: /D")
  short_cut <- dir_home |> dirname()
  system(paste0("subst x: \"", short_cut, "\""))

  dir_objects_names <- ls(pattern = "^dir_")
  dir_list <- mget(dir_objects_names)

  dir_list_shortened <- lapply(
    dir_list,
    function(y) create_shortcut(y, short_cut, "x")
  )
  list2env(dir_list_shortened, envir = .GlobalEnv)
}

rm(list = intersect(
  c("dir_objects_names", "short_cut", "dir_list", "dir_list_shortened"),
  ls()
))
```

# Parameters

```{r parameters}
PFAS_abbrev_names <- c("PFOS", "PFOA", "PFNA", "PFDA", "PFHxS", "PFHpS")
validation_cohort_levels <- c("SOLAR", "CHS", "SLP Age 25-35", "SLP Age 55-65", "BHRC")
validation_non_slp_cohort_levels <- c("SOLAR", "CHS", "BHRC")
validation_cohort_meta_levels <- c(validation_cohort_levels, "Meta-analysis")

wgcna_adjusted_p_threshold <- 0.20
wgcna_nominal_p_threshold <- 0.05

praxis_bf_threshold <- 1

dir_validation <- fs::path(dir_temp_data, "PFAS_final_validation")
fs::dir_create(dir_validation)
```

# Helper Functions

```{r workflow-source-helpers}
eval_qmd_until <- function(qmd_path, stop_pattern, envir = parent.frame()) {
  qmd_path <- here::here(qmd_path)
  r_file <- tempfile(fileext = ".R")
  knitr::purl(qmd_path, output = r_file, quiet = TRUE)

  r_lines <- readLines(r_file, warn = FALSE)
  stop_line <- grep(stop_pattern, r_lines)[1]

  if (is.na(stop_line)) {
    stop("Could not find stop pattern in tangled QMD: ", stop_pattern)
  }

  eval(parse(text = r_lines[seq_len(stop_line - 1)]), envir = envir)
}

collapse_unique <- function(x) {
  x <- unique(stats::na.omit(as.character(x)))
  paste(x, collapse = "; ")
}

make_effect_label <- function(beta, p_value) {
  dplyr::case_when(
    is.na(beta) ~ "",
    !is.na(p_value) & p_value < 0.01 ~ "**",
    !is.na(p_value) & p_value < 0.05 ~ "*",
    TRUE ~ ""
  )
}

inverse_variance_meta <- function(beta, se) {
  keep <- !is.na(beta) & !is.na(se) & se > 0

  if (sum(keep) == 0) {
    return(tibble::tibble(
      meta_n = 0L,
      estimate_beta = NA_real_,
      se_beta = NA_real_,
      ci_beta_low = NA_real_,
      ci_beta_high = NA_real_,
      p_value = NA_real_
    ))
  }

  weights <- 1 / se[keep]^2
  estimate <- sum(weights * beta[keep]) / sum(weights)
  se_meta <- sqrt(1 / sum(weights))
  z_value <- estimate / se_meta

  tibble::tibble(
    meta_n = sum(keep),
    estimate_beta = estimate,
    se_beta = se_meta,
    ci_beta_low = estimate - stats::qnorm(0.975) * se_meta,
    ci_beta_high = estimate + stats::qnorm(0.975) * se_meta,
    p_value = 2 * stats::pnorm(abs(z_value), lower.tail = FALSE)
  )
}

```

# WGCNA Review

This section reviews SOLAR reference WGCNA modules associated with PFAS in SOLAR, CHS, or BHRC projected-module analyses.
Modules are retained if any single PFAS or PFAS-mixture model has BH-adjusted p \< 0.20 or nominal p \< 0.05.
Hub proteins are selected using the same network inclusion rule used for the WGCNA circular network plots: top proteins by absolute module membership are filtered to proteins connected by TOM edges above the plot threshold.
The final hub list then keeps proteins labeled `Highly central` in the WGCNA circular network plot, which corresponds to the highest tertile of absolute module membership among network-included proteins.

```{r load-wgcna-review-objects}
eval_qmd_until(
  qmd_path = "1_scripts/!analytic_workflow_pfas_WGCNA.qmd",
  stop_pattern = "^SOLAR_module_circular_network_graphs <-"
)

wgcna_module_associations <- SOLAR_module_summary %>%
  dplyr::filter(
    !is.na(module_label),
    module != "grey",
    p.adjust.BH < wgcna_adjusted_p_threshold | p.value < wgcna_nominal_p_threshold
  ) %>%
  dplyr::arrange(p.adjust.BH, p.value)

selected_wgcna_modules <- wgcna_module_associations %>%
  dplyr::distinct(module_label, module, n_proteins) %>%
  dplyr::arrange(module_label)

selected_wgcna_modules %>%
  kableExtra::kbl(
    caption = paste0(
      "PFAS-associated SOLAR reference modules: BH-adjusted p < ",
      wgcna_adjusted_p_threshold,
      " or nominal p < ",
      wgcna_nominal_p_threshold,
      "."
    )
  ) %>%
  kableExtra::kable_paper("hover", full_width = FALSE)
```

```{r wgcna-hub-protein-selection}
summarise_module_graph_nodes <- function(module_label, module, n_proteins) {
  graph <- build_module_network(module, n_proteins = n_proteins)

  vertex_df <- igraph::as_data_frame(graph, what = "vertices") %>%
    tibble::as_tibble() %>%
    dplyr::rename(protein = name)

  edge_df <- igraph::as_data_frame(graph, what = "edges") %>%
    tibble::as_tibble()

  centrality_df <- dplyr::bind_rows(
    edge_df %>% dplyr::transmute(protein = from, TOM = TOM),
    edge_df %>% dplyr::transmute(protein = to, TOM = TOM)
  ) %>%
    dplyr::group_by(protein) %>%
    dplyr::summarise(
      TOM_degree = dplyr::n(),
      TOM_strength = sum(TOM, na.rm = TRUE),
      .groups = "drop"
    )

  vertex_df %>%
    dplyr::left_join(centrality_df, by = "protein") %>%
    dplyr::mutate(
      module_label = module_label,
      module = module,
      TOM_degree = dplyr::coalesce(TOM_degree, 0L),
      TOM_strength = dplyr::coalesce(TOM_strength, 0),
      kME_group_clean = stringr::str_replace_all(as.character(kME_group), "\\n", " ")
    ) %>%
    dplyr::relocate(module_label, module, protein, protein_label)
}

wgcna_network_nodes <- purrr::pmap_dfr(
  selected_wgcna_modules,
  summarise_module_graph_nodes
)

wgcna_hub_proteins <- wgcna_network_nodes %>%
  dplyr::filter(kME_group_clean == "Highly central") %>%
  dplyr::group_by(module_label, module) %>%
  dplyr::arrange(dplyr::desc(kME_abs), dplyr::desc(TOM_strength), dplyr::desc(TOM_degree), .by_group = TRUE) %>%
  dplyr::ungroup() %>%
  dplyr::left_join(
    wgcna_module_associations %>%
      dplyr::group_by(module_label, module) %>%
      dplyr::summarise(
        wgcna_associated_exposures = collapse_unique(exposure),
        wgcna_min_p = min(p.value, na.rm = TRUE),
        wgcna_min_BH = min(p.adjust.BH, na.rm = TRUE),
        .groups = "drop"
      ),
    by = c("module_label", "module")
  ) %>%
  dplyr::transmute(
    selection_source = "WGCNA hub",
    OlinkID = protein,
    Protein = protein_label,
    module_label,
    module,
    kME,
    kME_abs,
    TOM_degree,
    TOM_strength,
    wgcna_associated_exposures,
    wgcna_min_p,
    wgcna_min_BH
  )

wgcna_hub_proteins %>%
  dplyr::arrange(module_label, dplyr::desc(kME_abs)) %>%
  kableExtra::kbl(caption = "Highly central WGCNA hub proteins from PFAS-associated modules.") %>%
  kableExtra::kable_paper("hover", full_width = FALSE)
```

# Praxis-BGM Review

This section retrieves proteins with Praxis-BGM feature importance at or above 1 from the saved SOLAR reference and CHS/BHRC transfer models.
When a live Praxis-BGM Python model is available, the workflow uses the wrapper Bayes-factor selection helper.
If a model has been reloaded from RDS and the Python `BF_selection()` method is no longer callable, the workflow uses a saved-model fallback: standardized posterior cluster separation for each protein, calculated from the fitted posterior means and variances.
The fallback keeps the validation reproducible from saved model files and labels the metric explicitly as `posterior_cluster_separation`.

```{r load-praxis-review-objects}
eval_qmd_until(
  qmd_path = "1_scripts/!analytic_workflow_pfas_praxis_BGM.qmd",
  stop_pattern = "^set.seed\\(praxis_seed\\)"
)

if (!exists("praxis_bgm_bf_selection")) {
  praxis_r_interface_path <- Sys.getenv("PRAXIS_BGM_R_INTERFACE", unset = "")

  if (nzchar(praxis_r_interface_path) && fs::file_exists(praxis_r_interface_path)) {
    source(praxis_r_interface_path)
  } else {
    message(
      "Praxis-BGM R wrapper was not found locally; using posterior cluster ",
      "separation fallback for saved-model feature importance."
    )
  }
}

praxis_bf_file <- fs::path(dir_praxis, "Praxis_BGM_bayes_factor_top_features.csv")
SOLAR_praxis_model <- readRDS(fs::path(dir_praxis, "SOLAR_praxis_BGM_reference_model.rds"))
transfer_praxis_models <- readRDS(fs::path(dir_praxis, "CHS_BHRC_praxis_BGM_transfer_models.rds"))

praxis_feature_names <- SOLAR_praxis_model$feature_names

posterior_cluster_separation <- function(fit, cohort, feature_names) {
  mus <- fit$posterior_mus
  covs <- fit$posterior_covs

  if (is.null(mus) || is.null(covs) || length(dim(mus)) != 2 || length(dim(covs)) != 3) {
    stop("Saved Praxis-BGM fit does not contain posterior means and covariances.")
  }

  if (nrow(mus) != 2) {
    stop("Posterior separation fallback currently expects a two-cluster Praxis-BGM fit.")
  }

  pooled_var <- (diag(covs[1, , , drop = TRUE]) + diag(covs[2, , , drop = TRUE])) / 2
  pooled_sd <- sqrt(pmax(pooled_var, .Machine$double.eps))
  separation <- abs(mus[1, ] - mus[2, ]) / pooled_sd

  tibble::tibble(
    cohort = cohort,
    feature_index = seq_along(feature_names),
    OlinkID = feature_names,
    importance_value = as.numeric(separation),
    importance_metric = "posterior_cluster_separation"
  ) %>%
    dplyr::arrange(dplyr::desc(importance_value)) %>%
    dplyr::mutate(importance_rank = dplyr::row_number())
}

try_praxis_bf_selection <- function(model, data, cohort, feature_names) {
  if (!exists("praxis_bgm_bf_selection")) {
    return(posterior_cluster_separation(model, cohort, feature_names))
  }

  bf <- tryCatch(
    praxis_bgm_bf_selection(
      model = model,
      data = data,
      top_n = length(feature_names),
      visual = FALSE
    ),
    error = function(e) {
      message(
        "Praxis-BGM BF_selection() unavailable for ",
        cohort,
        "; using posterior cluster separation fallback. Original error: ",
        conditionMessage(e)
      )
      NULL
    }
  )

  if (is.null(bf)) {
    return(posterior_cluster_separation(model, cohort, feature_names))
  }

  out <- tibble::tibble(
    cohort = cohort,
    feature_index = bf$top_features,
    OlinkID = feature_names[bf$top_features],
    importance_value = NA_real_,
    importance_metric = "Praxis-BGM BF"
  )

  candidate_value_names <- c("bayes_factors", "bf_values", "BF", "scores", "feature_scores")
  value_name <- intersect(candidate_value_names, names(bf))[1]

  if (!is.na(value_name)) {
    values <- bf[[value_name]]
    if (length(values) == length(praxis_feature_names)) {
      out <- out %>%
        dplyr::mutate(importance_value = as.numeric(values[feature_index]))
    } else if (length(values) == length(bf$top_features)) {
      out <- out %>%
        dplyr::mutate(importance_value = as.numeric(values))
    }
  }

  if (all(is.na(out$importance_value))) {
    message(
      "Praxis-BGM BF_selection() returned no recognized BF/value field for ",
      cohort,
      "; using posterior cluster separation fallback."
    )
    return(posterior_cluster_separation(model, cohort, feature_names))
  }

  out %>%
    dplyr::arrange(dplyr::desc(importance_value)) %>%
    dplyr::mutate(importance_rank = dplyr::row_number())
}

if (fs::file_exists(praxis_bf_file)) {
  praxis_bf_top_features <- readr::read_csv(praxis_bf_file, show_col_types = FALSE) %>%
    dplyr::transmute(
      cohort,
      feature_index,
      OlinkID = feature_name,
      importance_value,
      importance_rank,
      importance_metric = "Praxis-BGM BF",
      Protein,
      UniProt
    )
} else {
  praxis_bf_top_features <- dplyr::bind_rows(
    try_praxis_bf_selection(SOLAR_praxis_model$fit, SOLAR_scaled$X, "SOLAR", praxis_feature_names),
    try_praxis_bf_selection(transfer_praxis_models$CHS$fit, CHS_scaled$X, "CHS", praxis_feature_names),
    try_praxis_bf_selection(transfer_praxis_models$BHRC$fit, BHRC_scaled$X, "BHRC", praxis_feature_names)
  ) %>%
    dplyr::left_join(
      proteomics_property %>%
        dplyr::select(OlinkID, Protein, UniProt),
      by = "OlinkID"
    )
}

praxis_top_proteins <- praxis_bf_top_features %>%
  dplyr::filter(!is.na(importance_value), importance_value >= praxis_bf_threshold) %>%
  dplyr::group_by(OlinkID, Protein, UniProt) %>%
  dplyr::summarise(
    selection_source = "Praxis-BGM importance",
    praxis_cohorts = collapse_unique(cohort),
    praxis_importance_metric = collapse_unique(importance_metric),
    praxis_max_importance = max(importance_value, na.rm = TRUE),
    praxis_best_rank = min(importance_rank, na.rm = TRUE),
    .groups = "drop"
  )

praxis_top_proteins %>%
  dplyr::arrange(dplyr::desc(praxis_max_importance), praxis_best_rank) %>%
  kableExtra::kbl(caption = paste0("Praxis-BGM proteins with feature importance >= ", praxis_bf_threshold, ".")) %>%
  kableExtra::kable_paper("hover", full_width = FALSE)
```

# Candidate Protein Pool

```{r pooled-candidate-proteins}
candidate_proteins <- dplyr::bind_rows(
  wgcna_hub_proteins %>%
    dplyr::select(selection_source, OlinkID, Protein, module_label, module, kME_abs, TOM_degree, TOM_strength, wgcna_associated_exposures, wgcna_min_p, wgcna_min_BH),
  praxis_top_proteins %>%
    dplyr::select(selection_source, OlinkID, Protein, praxis_cohorts, praxis_importance_metric, praxis_max_importance, praxis_best_rank)
) %>%
  dplyr::group_by(OlinkID) %>%
  dplyr::summarise(
    Protein = dplyr::coalesce(stats::na.omit(Protein)[1], OlinkID[1]),
    selection_sources = collapse_unique(selection_source),
    module_label = collapse_unique(module_label),
    module = collapse_unique(module),
    wgcna_associated_exposures = collapse_unique(wgcna_associated_exposures),
    wgcna_min_p = suppressWarnings(min(wgcna_min_p, na.rm = TRUE)),
    wgcna_min_BH = suppressWarnings(min(wgcna_min_BH, na.rm = TRUE)),
    max_kME_abs = suppressWarnings(max(kME_abs, na.rm = TRUE)),
    max_TOM_degree = suppressWarnings(max(TOM_degree, na.rm = TRUE)),
    max_TOM_strength = suppressWarnings(max(TOM_strength, na.rm = TRUE)),
    praxis_cohorts = collapse_unique(praxis_cohorts),
    praxis_importance_metric = collapse_unique(praxis_importance_metric),
    praxis_max_importance = suppressWarnings(max(praxis_max_importance, na.rm = TRUE)),
    praxis_best_rank = suppressWarnings(min(praxis_best_rank, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    dplyr::across(
      c(wgcna_min_p, wgcna_min_BH, max_kME_abs, max_TOM_degree, max_TOM_strength, praxis_max_importance, praxis_best_rank),
      ~ dplyr::if_else(is.infinite(.x), NA_real_, .x)
    ),
    display_label = dplyr::if_else(is.na(Protein) | Protein == "", OlinkID, Protein)
  ) %>%
  dplyr::arrange(selection_sources, module_label, dplyr::desc(praxis_max_importance), display_label)

readr::write_csv(
  candidate_proteins,
  fs::path(dir_validation, "PFAS_validation_candidate_proteins.csv")
)

candidate_proteins %>%
  kableExtra::kbl(caption = "Pooled WGCNA and Praxis-BGM candidate proteins.") %>%
  kableExtra::kable_paper("hover", full_width = FALSE)
```

# PWAS Effect Retrieval

```{r pwas-effect-retrieval}
pwas_single_file <- fs::path(dir_temp_data, "PFAS_PWAS", "PFAS_single_PWAS_results.csv")
slp_pwas_single_file <- fs::path(dir_temp_data, "PFAS_PWAS_BRI", "BRI_SLP_single_PFAS_PWAS_results.csv")

if (!fs::file_exists(pwas_single_file)) {
  stop("Missing SOLAR/CHS/BHRC single-PFAS PWAS file: ", pwas_single_file)
}

if (!fs::file_exists(slp_pwas_single_file)) {
  stop("Missing BRI SLP single-PFAS PWAS file: ", slp_pwas_single_file)
}

pwas_effects_main <- readr::read_csv(pwas_single_file, show_col_types = FALSE) %>%
  dplyr::filter(OlinkID %in% candidate_proteins$OlinkID) %>%
  dplyr::transmute(
    cohort = cohort,
    analysis_group = cohort,
    OlinkID,
    Protein,
    UniProt,
    pfas_abbrev,
    estimate_beta,
    se_beta,
    ci_beta_low,
    ci_beta_high,
    p_value,
    adjusted_pval_BH
  )

pwas_effects_slp <- readr::read_csv(slp_pwas_single_file, show_col_types = FALSE) %>%
  dplyr::filter(
    OlinkID %in% candidate_proteins$OlinkID,
    analysis_group %in% c("Age 25-35", "Age 55-65")
  ) %>%
  dplyr::transmute(
    cohort = "SLP",
    analysis_group = paste("SLP", analysis_group),
    OlinkID,
    Protein,
    UniProt,
    pfas_abbrev,
    estimate_beta,
    se_beta,
    ci_beta_low,
    ci_beta_high,
    p_value,
    adjusted_pval_BH
  )

pwas_candidate_effects <- dplyr::bind_rows(
  pwas_effects_main,
  pwas_effects_slp
) %>%
  dplyr::left_join(
    candidate_proteins %>%
      dplyr::select(OlinkID, selection_sources, module_label, wgcna_associated_exposures, praxis_max_importance, display_label),
    by = "OlinkID"
  ) %>%
  dplyr::mutate(
    analysis_group = factor(analysis_group, levels = validation_cohort_levels),
    pfas_abbrev = factor(pfas_abbrev, levels = PFAS_abbrev_names),
    estimate_beta_log2 = estimate_beta,
    effect_label = make_effect_label(estimate_beta_log2, p_value)
  ) %>%
  dplyr::arrange(display_label, analysis_group, pfas_abbrev)

readr::write_csv(
  pwas_candidate_effects,
  fs::path(dir_validation, "PFAS_validation_candidate_PWAS_effects_long.csv")
)

pwas_candidate_effects_wide <- pwas_candidate_effects %>%
  dplyr::select(
    display_label,
    OlinkID,
    Protein,
    selection_sources,
    analysis_group,
    pfas_abbrev,
    estimate_beta_log2,
    p_value,
    adjusted_pval_BH
  ) %>%
  tidyr::pivot_wider(
    names_from = c(analysis_group, pfas_abbrev),
    values_from = c(estimate_beta_log2, p_value, adjusted_pval_BH),
    names_glue = "{analysis_group}_{pfas_abbrev}_{.value}"
  )

readr::write_csv(
  pwas_candidate_effects_wide,
  fs::path(dir_validation, "PFAS_validation_candidate_PWAS_effects_wide.csv")
)

pwas_candidate_effects_wide %>%
  kableExtra::kbl(caption = "Candidate-protein PWAS effect estimates across cohorts and SLP age groups.") %>%
  kableExtra::kable_paper("hover", full_width = FALSE)
```

# PFAS-Protein Meta-analysis

For each prioritized protein and PFAS pair, cohort-specific PWAS estimates are combined using fixed-effect inverse-variance meta-analysis.
The meta-estimate is on the same per-doubling PFAS scale as the cohort-specific PWAS estimates.

```{r pwas-effect-meta-analysis}
pwas_candidate_meta_effects <- pwas_candidate_effects %>%
  dplyr::filter(!is.na(analysis_group)) %>%
  dplyr::group_by(OlinkID, Protein, UniProt, pfas_abbrev) %>%
  dplyr::group_modify(~ inverse_variance_meta(.x$estimate_beta_log2, .x$se_beta)) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    adjusted_pval_BH = stats::p.adjust(p_value, method = "BH"),
    cohort = "Meta-analysis",
    analysis_group = factor("Meta-analysis", levels = validation_cohort_meta_levels),
    pfas_abbrev = factor(pfas_abbrev, levels = PFAS_abbrev_names),
    estimate_beta_log2 = estimate_beta,
    effect_label = make_effect_label(estimate_beta_log2, p_value)
  ) %>%
  dplyr::left_join(
    candidate_proteins %>%
      dplyr::select(OlinkID, selection_sources, module_label, wgcna_associated_exposures, praxis_max_importance, display_label),
    by = "OlinkID"
  ) %>%
  dplyr::select(
    cohort,
    analysis_group,
    OlinkID,
    Protein,
    UniProt,
    pfas_abbrev,
    meta_n,
    estimate_beta,
    se_beta,
    ci_beta_low,
    ci_beta_high,
    p_value,
    adjusted_pval_BH,
    selection_sources,
    module_label,
    wgcna_associated_exposures,
    praxis_max_importance,
    display_label,
    estimate_beta_log2,
    effect_label
  ) %>%
  dplyr::arrange(display_label, pfas_abbrev)

readr::write_csv(
  pwas_candidate_meta_effects,
  fs::path(dir_validation, "PFAS_validation_candidate_PWAS_meta_effects.csv")
)

pwas_candidate_meta_effects %>%
  dplyr::select(
    display_label,
    OlinkID,
    Protein,
    pfas_abbrev,
    meta_n,
    estimate_beta_log2,
    se_beta,
    ci_beta_low,
    ci_beta_high,
    p_value,
    adjusted_pval_BH
  ) %>%
  kableExtra::kbl(caption = "Fixed-effect inverse-variance meta-analysis of prioritized PFAS-protein associations.") %>%
  kableExtra::kable_paper("hover", full_width = FALSE)
```

# Effect Estimate Heatmap

The heatmap focuses on candidate proteins with PWAS estimates available in all validation analysis groups: SOLAR, CHS, BHRC, SLP Age 25-35, and SLP Age 55-65.
Because the PWAS models were fit using log2-transformed PFAS exposures, the fitted beta is the NPX change per one-unit increase in log2 PFAS, equivalent to a doubling in PFAS concentration.
The heatmap therefore displays the original PWAS effect estimates on the per-doubling PFAS scale.

```{r pwas-effect-heatmap, fig.width=12, fig.height=8}
overlap_candidate_proteins <- pwas_candidate_effects %>%
  dplyr::distinct(OlinkID, analysis_group) %>%
  dplyr::count(OlinkID, name = "n_analysis_groups") %>%
  dplyr::filter(n_analysis_groups == length(validation_cohort_levels)) %>%
  dplyr::inner_join(candidate_proteins, by = "OlinkID")

if (nrow(overlap_candidate_proteins) == 0) {
  stop("No candidate proteins have PWAS estimates in all validation analysis groups.")
}

protein_order <- pwas_candidate_effects %>%
  dplyr::filter(OlinkID %in% overlap_candidate_proteins$OlinkID) %>%
  dplyr::distinct(display_label, selection_sources, module_label, praxis_max_importance) %>%
  dplyr::arrange(selection_sources, module_label, dplyr::desc(praxis_max_importance), display_label) %>%
  dplyr::pull(display_label)

pwas_effect_heatmap_df <- pwas_candidate_effects %>%
  dplyr::filter(OlinkID %in% overlap_candidate_proteins$OlinkID) %>%
  dplyr::mutate(
    display_label = factor(display_label, levels = rev(unique(protein_order))),
    analysis_group = factor(analysis_group, levels = validation_cohort_levels),
    pfas_abbrev = factor(pfas_abbrev, levels = PFAS_abbrev_names),
    sig_label = dplyr::case_when(
      !is.na(p_value) & p_value < 0.01 ~ "**",
      !is.na(p_value) & p_value < 0.05 ~ "*",
      TRUE ~ ""
    )
  )

pwas_effect_heatmap <- ggplot2::ggplot(
  pwas_effect_heatmap_df,
  ggplot2::aes(x = analysis_group, y = display_label, fill = estimate_beta_log2)
) +
  ggplot2::geom_tile(color = "white", linewidth = 0.25) +
  ggplot2::geom_text(ggplot2::aes(label = sig_label), size = 2.6) +
  ggplot2::facet_grid(. ~ pfas_abbrev, drop = FALSE) +
  ggplot2::scale_fill_gradient2(
    low = "#2C7BB6",
    mid = "grey98",
    high = "#D7191C",
    midpoint = 0,
    na.value = "grey90",
    name = "NPX change per PFAS doubling"
  ) +
  ggplot2::labs(
    x = NULL,
    y = NULL,
    # caption = paste(
    #   "Only candidate proteins with PWAS estimates in SOLAR, CHS, BHRC, and both SLP age groups are shown.",
    #   "Effect estimates are original PWAS betas: NPX change per doubling in PFAS concentration.",
    #   "* nominal p < 0.05; ** nominal p < 0.01"
    # )
  ) +
  ggplot2::theme_bw(base_size = 10) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1, vjust = 1),
    panel.grid = ggplot2::element_blank(),
    strip.background = ggplot2::element_rect(fill = "grey90", color = NA),
    legend.position = "bottom",
    plot.caption = ggplot2::element_text(hjust = 0)
  )

pwas_effect_heatmap

ggplot2::ggsave(
  filename = fs::path(dir_figures, "PFAS_final_validation_candidate_PWAS_heatmap.png"),
  plot = pwas_effect_heatmap,
  dpi = 600,
  units = "in",
  width = 6.5,
  height = 4.5
)
```

## Effect Estimate Heatmap With Meta-analysis

This heatmap uses the same prioritized proteins and cohort-specific PWAS estimates as the all-cohorts heatmap, with an added fixed-effect meta-analysis column for each PFAS.

```{r pwas-effect-heatmap-with-meta-analysis, fig.width=13, fig.height=8}
pwas_effect_heatmap_with_meta_df <- dplyr::bind_rows(
  pwas_effect_heatmap_df %>%
    dplyr::mutate(
      analysis_group = as.character(analysis_group),
      pfas_abbrev = as.character(pfas_abbrev)
    ),
  pwas_candidate_meta_effects %>%
    dplyr::filter(OlinkID %in% overlap_candidate_proteins$OlinkID) %>%
    dplyr::mutate(
      sig_label = dplyr::case_when(
        !is.na(p_value) & p_value < 0.01 ~ "**",
        !is.na(p_value) & p_value < 0.05 ~ "*",
        TRUE ~ ""
      ),
      analysis_group = as.character(analysis_group),
      pfas_abbrev = as.character(pfas_abbrev)
    )
) %>%
  dplyr::mutate(
    display_label = factor(as.character(display_label), levels = rev(unique(protein_order))),
    analysis_group = factor(analysis_group, levels = validation_cohort_meta_levels),
    pfas_abbrev = factor(pfas_abbrev, levels = PFAS_abbrev_names)
  )

pwas_effect_heatmap_with_meta <- ggplot2::ggplot(
  pwas_effect_heatmap_with_meta_df,
  ggplot2::aes(x = analysis_group, y = display_label, fill = estimate_beta_log2)
) +
  ggplot2::geom_tile(color = "white", linewidth = 0.25) +
  ggplot2::geom_text(ggplot2::aes(label = sig_label), size = 2.6) +
  ggplot2::facet_grid(. ~ pfas_abbrev, drop = FALSE) +
  ggplot2::scale_fill_gradient2(
    low = "#2C7BB6",
    mid = "grey98",
    high = "#D7191C",
    midpoint = 0,
    na.value = "grey90",
    name = "NPX change per PFAS doubling"
  ) +
  ggplot2::labs(
    x = NULL,
    y = NULL
  ) +
  ggplot2::theme_bw(base_size = 10) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1, vjust = 1),
    panel.grid = ggplot2::element_blank(),
    strip.background = ggplot2::element_rect(fill = "grey90", color = NA),
    legend.position = "bottom",
    plot.caption = ggplot2::element_text(hjust = 0)
  )

pwas_effect_heatmap_with_meta

ggplot2::ggsave(
  filename = fs::path(dir_figures, "PFAS_final_validation_candidate_PWAS_heatmap_with_meta_analysis.png"),
  plot = pwas_effect_heatmap_with_meta,
  dpi = 600,
  units = "in",
  width = 7.5,
  height = 4.5
)
```

## Effect Estimate Heatmap Table

This table includes the PWAS effect estimates and nominal p-values displayed in the all-cohorts heatmap.

```{r pwas-effect-heatmap-table}
pwas_effect_heatmap_estimates_nominal_p <- pwas_effect_heatmap_df %>%
  dplyr::mutate(
    display_label = as.character(display_label),
    analysis_group = as.character(analysis_group),
    pfas_abbrev = as.character(pfas_abbrev),
    estimate_beta_log2 = dplyr::case_when(
      is.na(estimate_beta_log2) ~ NA_character_,
      !is.na(p_value) & p_value < 0.01 ~ paste0(sprintf("%.2f", estimate_beta_log2), "**"),
      !is.na(p_value) & p_value < 0.05 ~ paste0(sprintf("%.2f", estimate_beta_log2), "*"),
      TRUE ~ sprintf("%.2f", estimate_beta_log2)
    )
  ) %>%
  dplyr::select(
    OlinkID,
    Protein,
    display_label,
    selection_sources,
    analysis_group,
    pfas_abbrev,
    estimate_beta_log2
  ) %>%
  tidyr::pivot_wider(
    names_from = c(analysis_group, pfas_abbrev),
    values_from = estimate_beta_log2,
    names_glue = "{analysis_group}_{pfas_abbrev}_{.value}"
  )

# readr::write_csv(
#   pwas_effect_heatmap_estimates_nominal_p,
#   fs::path(dir_validation, "PFAS_final_validation_candidate_PWAS_heatmap_all_cohorts_estimates_nominal_p.csv")
# )

pwas_effect_heatmap_estimates_nominal_p %>%
  kableExtra::kbl(
    caption = "Candidate-protein PWAS effect estimates across all validation cohorts"
  ) %>%
  kableExtra::kable_paper("hover", full_width = FALSE) %>%
  kableExtra::footnote(
    general = c(
      "Values are PWAS beta estimates representing NPX change per doubling in PFAS concentration.",
      "Asterisks indicate nominal association strength: * p < 0.05; ** p < 0.01.",
      "The table includes the candidate proteins displayed in the all-cohorts heatmap."
    ),
    general_title = "Notes: ",
    footnote_as_chunk = TRUE
  )
```

## Effect Estimate Heatmap Without SLP Cohorts

This heatmap focuses on candidate proteins with PWAS estimates available in the non-SLP validation cohorts: SOLAR, CHS, and BHRC.
The heatmap displays the original PWAS effect estimates on the per-doubling PFAS scale.

```{r pwas-effect-heatmap-no-slp, fig.width=12, fig.height=8}
overlap_candidate_proteins_no_slp <- pwas_candidate_effects %>%
  dplyr::filter(analysis_group %in% validation_non_slp_cohort_levels) %>%
  dplyr::distinct(OlinkID, analysis_group) %>%
  dplyr::count(OlinkID, name = "n_analysis_groups") %>%
  dplyr::filter(n_analysis_groups == length(validation_non_slp_cohort_levels)) %>%
  dplyr::inner_join(candidate_proteins, by = "OlinkID")

if (nrow(overlap_candidate_proteins_no_slp) == 0) {
  stop("No candidate proteins have PWAS estimates in all non-SLP validation cohorts.")
}

protein_order_no_slp <- pwas_candidate_effects %>%
  dplyr::filter(OlinkID %in% overlap_candidate_proteins_no_slp$OlinkID) %>%
  dplyr::distinct(display_label, selection_sources, module_label, praxis_max_importance) %>%
  dplyr::arrange(selection_sources, module_label, dplyr::desc(praxis_max_importance), display_label) %>%
  dplyr::pull(display_label)

pwas_effect_heatmap_no_slp_df <- pwas_candidate_effects %>%
  dplyr::filter(
    OlinkID %in% overlap_candidate_proteins_no_slp$OlinkID,
    analysis_group %in% validation_non_slp_cohort_levels
  ) %>%
  dplyr::mutate(
    display_label = factor(display_label, levels = rev(unique(protein_order_no_slp))),
    analysis_group = factor(analysis_group, levels = validation_non_slp_cohort_levels),
    pfas_abbrev = factor(pfas_abbrev, levels = PFAS_abbrev_names),
    sig_label = dplyr::case_when(
      !is.na(p_value) & p_value < 0.01 ~ "**",
      !is.na(p_value) & p_value < 0.05 ~ "*",
      TRUE ~ ""
    )
  )

pwas_effect_heatmap_no_slp <- ggplot2::ggplot(
  pwas_effect_heatmap_no_slp_df,
  ggplot2::aes(x = display_label, y = analysis_group, fill = estimate_beta_log2)
) +
  ggplot2::geom_tile(color = "white", linewidth = 0.25) +
  ggplot2::geom_text(ggplot2::aes(label = sig_label), size = 2.6) +
  ggplot2::facet_grid(pfas_abbrev ~ ., drop = FALSE) +
  ggplot2::scale_fill_gradient2(
    low = "#2C7BB6",
    mid = "grey98",
    high = "#D7191C",
    midpoint = 0,
    na.value = "grey90",
    name = "NPX change per PFAS doubling"
  ) +
  ggplot2::labs(
    x = NULL,
    y = NULL
  ) +
  ggplot2::theme_bw(base_size = 10) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1, vjust = 1),
    panel.grid = ggplot2::element_blank(),
    strip.background = ggplot2::element_rect(fill = "grey90", color = NA),
    strip.text.y = ggplot2::element_text(angle = 0),
    legend.position = "bottom",
    plot.caption = ggplot2::element_text(hjust = 0)
  )

pwas_effect_heatmap_no_slp

ggplot2::ggsave(
  filename = fs::path(dir_figures, "PFAS_final_validation_candidate_PWAS_heatmap_no_SLP.png"),
  plot = pwas_effect_heatmap_no_slp,
  dpi = 600,
  units = "in",
  width = 8,
  height = 6
)
```

# Supplementary Data 1 Export

This optional export creates an Excel workbook with one sheet for each cohort or SLP age group and one sheet for the fixed-effect meta-analysis results.

```{r supplementary-data-1-export}
#| eval: false

if (!requireNamespace("openxlsx", quietly = TRUE)) {
  stop("Package 'openxlsx' is required to write Supplementary Data 1.")
}

supplementary_data_1_file <- fs::path(
  dir_temp_data,
  "Supplementary_Data_1_PFAS_protein_associations.xlsx"
)

supplementary_data_1_cohort_sheets <- validation_cohort_levels %>%
  rlang::set_names() %>%
  purrr::map(~ pwas_candidate_effects %>%
    dplyr::filter(as.character(analysis_group) == .x) %>%
    dplyr::mutate(
      analysis_group = as.character(analysis_group),
      pfas_abbrev = as.character(pfas_abbrev)
    ) %>%
    dplyr::arrange(display_label, pfas_abbrev)
  )

supplementary_data_1_sheets <- c(
  supplementary_data_1_cohort_sheets,
  list(
    `Meta-analysis` = pwas_candidate_meta_effects %>%
      dplyr::mutate(
        analysis_group = as.character(analysis_group),
        pfas_abbrev = as.character(pfas_abbrev)
      )
  )
)

supplementary_data_1_sheets <- supplementary_data_1_sheets %>%
  purrr::map(~ .x %>%
    dplyr::arrange(display_label, pfas_abbrev) %>%
    dplyr::select(
      cohort,
      OlinkID,
      Protein,
      UniProt,
      pfas_abbrev,
      estimate_beta_log2,
      se_beta,
      ci_beta_low,
      ci_beta_high,
      p_value,
      adjusted_pval_BH,
      tidyselect::any_of("meta_n"),
      selection_sources
    )
  )

openxlsx::write.xlsx(
  supplementary_data_1_sheets,
  file = supplementary_data_1_file,
  overwrite = TRUE
)
```
